---
title: "Fiscal Policy by Administration"
subtitle: "Federal deficit and national debt under each president"
---

```{r setup}
#| include: false
source("R/helpers.R")
source("R/presidential_helpers.R")

admins   <- load_admin_summary()
annual   <- load_annual()
fiscal_q <- load_fiscal_quarterly() |> mutate(date = as.Date(date))

admins <- admins |>
  mutate(
    label       = admin_short_label(president, ongoing),
    party_color = party_color(party)
  )

color_array_js <- sprintf("[%s]",
  paste(sprintf('"%s"', admins$party_color), collapse = ","))

bar_color_js <- htmlwidgets::JS(sprintf(
  "function(p){ var c=%s; return c[p.dataIndex]; }", color_array_js))

# Colored bands per admin — used on timeline charts below
bands <- admins |>
  transmute(start = format(window_first, "%Y-%m-%d"),
            end   = format(window_last,  "%Y-%m-%d"),
            color = party_color, label)
mark_data <- lapply(seq_len(nrow(bands)), function(i) list(
  list(xAxis = bands$start[i],
       itemStyle = list(color = paste0(bands$color[i], "22")),
       name = bands$label[i]),
  list(xAxis = bands$end[i])
))
```

::: {.callout-important appearance="simple"}
Attribution caveats compound on fiscal data. The president who takes office
in January **inherits** the fiscal year that started the previous October,
which was drafted and enacted under the prior administration. Major
legislation (tax cuts, stimulus, entitlement reform) is passed by Congress,
not the White House. Recessions and wars drive deficits regardless of who
holds office. Read these numbers as *descriptive context* for each
administration's fiscal environment, not as a causal scorecard.
See [Methodology](about.qmd) for more.
:::

## Federal debt — total outstanding {#debt-level}

Quarterly level of federal debt held by the public, shaded by administration.

```{r debt-level}
#| echo: false
fiscal_q |>
  e_charts(date) |>
  e_line(debt_trillion, name = "Federal debt ($T)", symbol = "none", smooth = TRUE,
         lineStyle = list(width = 2, color = "#1d3557")) |>
  e_mark_area(data = mark_data, silent = TRUE,
              label = list(show = TRUE, position = "insideTop",
                           color = "#495057", fontSize = 10)) |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? '$'+v.toFixed(2)+' T' : v; }")) |>
  e_datazoom(x_index = 0) |>
  e_y_axis(name = "Trillions of $") |>
  e_title("Federal debt outstanding — shaded by administration",
          left = "center") |>
  apply_econ_theme()
```

## Federal debt as % of GDP {#debt-pct-gdp}

Debt-to-GDP normalizes for economic growth and inflation — a better
long-run stress metric than the nominal dollar amount.

```{r debt-pct-gdp}
#| echo: false
fiscal_q |>
  e_charts(date) |>
  e_line(debt_pct_gdp, name = "Debt / GDP (%)", symbol = "none", smooth = TRUE,
         lineStyle = list(width = 2, color = "#6a4c93")) |>
  e_mark_area(data = mark_data, silent = TRUE,
              label = list(show = TRUE, position = "insideTop",
                           color = "#495057", fontSize = 10)) |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? v.toFixed(1)+'%' : v; }")) |>
  e_datazoom(x_index = 0) |>
  e_y_axis(name = "%", axisLabel = list(formatter = "{value}%")) |>
  e_title("Federal debt as a share of GDP", left = "center") |>
  apply_econ_theme()
```

## Annual deficit (or surplus) as % of GDP {#deficit-annual}

Positive values = surplus, negative = deficit. Bars colored by the
administration in office the year they appeared (mid-year rule).

```{r deficit-bars}
#| echo: false
source("R/presidential_helpers.R")

# Tag each calendar year with the admin in office on July 1
admins_raw <- load_administrations()
tag_admin <- function(d) {
  idx <- which(d >= admins_raw$start_date & d < admins_raw$effective_end)
  if (length(idx) == 0) NA_integer_ else idx[1]
}

deficit_years <- annual |>
  filter(!is.na(deficit_pct_gdp)) |>
  mutate(
    mid = as.Date(sprintf("%d-07-01", year)),
    admin_idx = vapply(mid, tag_admin, integer(1)),
    party = ifelse(is.na(admin_idx), "Unknown", admins_raw$party[admin_idx]),
    color = ifelse(is.na(admin_idx), "#adb5bd", party_color(party)),
    label = ifelse(prototype == 1, paste0(year, "*"), as.character(year))
  )

bar_color_year <- htmlwidgets::JS(sprintf(
  "function(p){ var c=[%s]; return c[p.dataIndex]; }",
  paste(sprintf('"%s"', deficit_years$color), collapse = ",")))

deficit_years |>
  e_charts(label) |>
  e_bar(deficit_pct_gdp, name = "Deficit / surplus (% GDP)",
        itemStyle = list(color = bar_color_year)) |>
  e_mark_line(data = list(yAxis = 0), lineStyle = list(color = "#6c757d"),
              symbol = "none") |>
  e_x_axis(axisLabel = list(rotate = 30, interval = 0)) |>
  e_y_axis(name = "% of GDP", axisLabel = list(formatter = "{value}%")) |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? v.toFixed(1)+'%' : v; }")) |>
  e_title("Federal surplus / deficit as % of GDP",
          subtext = "Positive = surplus; negative = deficit. Bars colored by party in office mid-year",
          left = "center") |>
  apply_econ_theme()
```

## Debt added during each administration {#debt-added}

```{r debt-added}
#| echo: false
admins |>
  e_charts(label) |>
  e_bar(debt_added_trillion, name = "Debt added ($T)",
        itemStyle = list(color = bar_color_js)) |>
  e_mark_line(data = list(yAxis = 0), lineStyle = list(color = "#6c757d"),
              symbol = "none") |>
  e_y_axis(name = "Trillions of $") |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? '$'+v.toFixed(2)+' T' : v; }")) |>
  e_title("Increase in federal debt during each administration",
          subtext = "End-of-term debt minus first-quarter-in-window debt",
          left = "center") |>
  apply_econ_theme()
```

## Average deficit as % of GDP {#avg-deficit}

```{r avg-deficit}
#| echo: false
admins |>
  e_charts(label) |>
  e_bar(avg_deficit_pct_gdp, name = "Avg deficit / surplus (% GDP)",
        itemStyle = list(color = bar_color_js)) |>
  e_mark_line(data = list(yAxis = 0), lineStyle = list(color = "#6c757d"),
              symbol = "none") |>
  e_y_axis(name = "% of GDP", axisLabel = list(formatter = "{value}%")) |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? v.toFixed(2)+'%' : v; }")) |>
  e_title("Average annual deficit / surplus as % of GDP, per administration",
          left = "center") |>
  apply_econ_theme()
```

## Fiscal summary by administration {#fiscal-table}

```{r fiscal-table}
#| echo: false
library(reactable)
library(htmltools)

party_badge <- function(party) {
  col <- party_color(party)
  tags$span(
    style = sprintf(
      "display:inline-block; padding:2px 8px; border-radius:3px; font-size:0.72rem;
       font-weight:600; color:#fff; background:%s;", col),
    toupper(party)
  )
}

admins |>
  transmute(
    president = paste0(president, ifelse(ongoing, " *", "")),
    party,
    debt_start_trillion,
    debt_end_trillion,
    debt_added_trillion,
    debt_pct_gdp_start,
    debt_pct_gdp_end,
    debt_pct_gdp_change,
    avg_deficit_pct_gdp
  ) |>
  reactable(
    striped = TRUE, highlight = TRUE, compact = TRUE,
    theme = reactableTheme(borderColor = "#dee2e6"),
    columns = list(
      president = colDef(name = "President", minWidth = 180),
      party = colDef(name = "Party", minWidth = 110,
                     cell = function(v) party_badge(v), html = TRUE),
      debt_start_trillion = colDef(name = "Debt start ($T)",
                                   format = colFormat(digits = 2)),
      debt_end_trillion = colDef(name = "Debt end ($T)",
                                 format = colFormat(digits = 2)),
      debt_added_trillion = colDef(name = "Debt added ($T)",
                                   format = colFormat(digits = 2),
                                   style = list(fontWeight = 600)),
      debt_pct_gdp_start = colDef(name = "Debt/GDP start",
                                  format = colFormat(digits = 1, suffix = "%")),
      debt_pct_gdp_end = colDef(name = "Debt/GDP end",
                                format = colFormat(digits = 1, suffix = "%")),
      debt_pct_gdp_change = colDef(name = "Δ Debt/GDP (pp)",
                                   format = colFormat(digits = 1)),
      avg_deficit_pct_gdp = colDef(name = "Avg deficit (% GDP)",
                                   format = colFormat(digits = 2, suffix = "%"),
                                   style = function(v) list(
                                     color = ifelse(!is.na(v) & v < 0, "#bc4749", "#2f9e44"),
                                     fontWeight = 600))
    )
  )
```
