---
title: "Growth by Administration"
subtitle: "Real GDP, unemployment, and inflation under each president"
---

```{r setup}
#| include: false
source("R/helpers.R")
source("R/presidential_helpers.R")

admins  <- load_admin_summary()
monthly <- load_monthly_admin() |> mutate(date = as.Date(date))

admins <- admins |>
  mutate(
    label       = admin_short_label(president, ongoing),
    party_color = party_color(party)
  )

color_array_js <- sprintf("[%s]",
  paste(sprintf('"%s"', admins$party_color), collapse = ","))

bar_color_js <- htmlwidgets::JS(sprintf(
  "function(p){ var c=%s; return c[p.dataIndex]; }", color_array_js))
```

::: {.callout-note appearance="simple"}
All comparisons are descriptive, not causal. See [Methodology](about.qmd).
:::

## Real GDP growth {#gdp}

```{r gdp}
#| echo: false
admins |>
  e_charts(label) |>
  e_bar(gdp_growth_avg, name = "Avg annual GDP growth (%)",
        itemStyle = list(color = bar_color_js)) |>
  e_y_axis(name = "%", axisLabel = list(formatter = "{value}%")) |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? v.toFixed(2)+'%' : v; }")) |>
  e_title("Average real GDP growth per administration",
          subtext = "Calendar-year GDP growth, averaged across years in admin window",
          left = "center") |>
  apply_econ_theme()
```

## Unemployment — start vs end {#unemployment}

Bars show the unemployment rate at the admin's first full month vs. last
month in the data window. Downward movement means unemployment fell.

```{r unemployment}
#| echo: false
u <- admins |>
  transmute(label, Start = unemployment_start, End = unemployment_end,
            party_color)

u |>
  e_charts(label) |>
  e_bar(Start, name = "First month",
        itemStyle = list(color = "#adb5bd")) |>
  e_bar(End,   name = "Last month",
        itemStyle = list(color = bar_color_js)) |>
  e_y_axis(name = "%", axisLabel = list(formatter = "{value}%")) |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? v.toFixed(2)+'%' : v; }")) |>
  e_title("Unemployment rate — first vs last month per administration",
          left = "center") |>
  e_legend(top = 56) |>
  apply_econ_theme()
```

## Inflation (CPI, year-over-year) {#cpi}

```{r cpi}
#| echo: false
admins |>
  e_charts(label) |>
  e_bar(cpi_avg, name = "Avg CPI inflation (%)",
        itemStyle = list(color = bar_color_js)) |>
  e_y_axis(name = "%", axisLabel = list(formatter = "{value}%")) |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? v.toFixed(2)+'%' : v; }")) |>
  e_title("Average CPI inflation per administration", left = "center") |>
  apply_econ_theme()
```

## Monthly unemployment, administration-shaded {#unemployment-timeline}

```{r unemployment-line}
#| echo: false
bands <- admins |>
  transmute(start = format(window_first, "%Y-%m-%d"),
            end   = format(window_last,  "%Y-%m-%d"),
            color = party_color, label)
mark_data <- lapply(seq_len(nrow(bands)), function(i) list(
  list(xAxis = bands$start[i],
       itemStyle = list(color = paste0(bands$color[i], "22")),
       name = bands$label[i]),
  list(xAxis = bands$end[i])
))

monthly |>
  e_charts(date) |>
  e_line(unemployment, name = "Unemployment (%)", symbol = "none", smooth = TRUE,
         lineStyle = list(width = 2, color = "#1d3557")) |>
  e_mark_area(data = mark_data, silent = TRUE,
              label = list(show = TRUE, position = "insideTop",
                           color = "#495057", fontSize = 10)) |>
  e_tooltip(trigger = "axis") |>
  e_datazoom(x_index = 0) |>
  e_y_axis(name = "%", axisLabel = list(formatter = "{value}%")) |>
  e_title("Unemployment rate — by administration", left = "center") |>
  apply_econ_theme()
```
