---
title: "Presidential Economies"
subtitle: "Economic indicators and markets by administration, 1999 to present"
---

```{r setup}
#| include: false
source("R/helpers.R")
source("R/presidential_helpers.R")

admins  <- load_admin_summary()
monthly <- load_monthly_admin() |> mutate(date = as.Date(date))

admins <- admins |>
  mutate(
    label   = admin_short_label(president, ongoing),
    party_color = party_color(party)
  )
```

This report slices the same 1999-to-present US economic data used in the
[US Economy](../economy/index.qmd) analysis by the president in office. Each
calendar month is assigned to whichever president held office on the 14th of
that month; inaugurations on January 20 mean the incoming president's first
*full* month is February.

::: {.callout-important appearance="simple"}
**Read the caveats before drawing conclusions.** Presidents do not fully
control the economy. Inherited conditions, Federal Reserve monetary policy,
external shocks (dot-com bust, 2008 financial crisis, COVID-19), and
multi-year policy lags all dwarf the effect of any single White House.
See the [methodology page](about.qmd) for a fuller discussion.
:::

Administrations marked with an asterisk (`*`) are ongoing — their stats
cover only the portion of their term observed so far.

## Summary table {#summary}

```{r summary-table}
#| echo: false
library(reactable)
library(htmltools)

party_badge <- function(party) {
  col <- party_color(party)
  htmltools::tags$span(
    style = sprintf(
      "display:inline-block; padding:2px 8px; border-radius:3px; font-size:0.72rem;
       font-weight:600; letter-spacing:0.04em; color:#fff; background:%s;", col),
    toupper(party)
  )
}

tbl <- admins |>
  transmute(
    president = paste0(president, ifelse(ongoing, " *", "")),
    party,
    window = sprintf("%s – %s", format(window_first, "%b %Y"),
                                format(window_last,  "%b %Y")),
    months,
    gdp_growth_avg,
    unemployment_change,
    cpi_avg,
    sp500_total_return,
    sp500_annualized_return,
    vix_avg
  )

reactable(
  tbl,
  striped = TRUE, highlight = TRUE, compact = TRUE,
  theme = reactableTheme(borderColor = "#dee2e6"),
  columns = list(
    president = colDef(name = "President", minWidth = 180),
    party = colDef(name = "Party", minWidth = 110,
                   cell = function(v) party_badge(v), html = TRUE),
    window = colDef(name = "Window", minWidth = 170),
    months = colDef(name = "Months", maxWidth = 80),
    gdp_growth_avg = colDef(name = "GDP avg (%)",
                            format = colFormat(digits = 2, suffix = "%")),
    unemployment_change = colDef(name = "ΔUnemployment (pp)",
                                 format = colFormat(digits = 1)),
    cpi_avg = colDef(name = "CPI avg (%)",
                     format = colFormat(digits = 2, suffix = "%")),
    sp500_total_return = colDef(name = "S&P total (%)",
                                format = colFormat(digits = 1, suffix = "%"),
                                style = function(v) list(
                                  fontWeight = 600,
                                  color = ifelse(v < 0, "#bc4749", "#2f9e44"))),
    sp500_annualized_return = colDef(name = "S&P annual (%)",
                                     format = colFormat(digits = 1, suffix = "%")),
    vix_avg = colDef(name = "VIX avg")
  )
)
```

## GDP growth by administration {#gdp}

```{r gdp-bar}
#| echo: false
admins |>
  e_charts(label) |>
  e_bar(gdp_growth_avg, name = "Avg GDP growth (%)",
        itemStyle = list(
          color = htmlwidgets::JS(sprintf("
            function(p){
              var colors = [%s];
              return colors[p.dataIndex];
            }",
            paste(sprintf('\"%s\"', admins$party_color), collapse=",")
          ))
        )) |>
  e_y_axis(name = "%", axisLabel = list(formatter = "{value}%")) |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v === 'number' ? v.toFixed(2) + '%' : v; }")) |>
  e_title("Average real GDP growth during each administration",
          subtext = "Annual GDP growth, calendar years averaged by admin",
          left = "center") |>
  apply_econ_theme()
```

## S&P 500 total return by administration {#sp500}

```{r sp-bar}
#| echo: false
admins |>
  e_charts(label) |>
  e_bar(sp500_total_return, name = "Total return (%)",
        itemStyle = list(
          color = htmlwidgets::JS(sprintf("
            function(p){
              var colors = [%s];
              return colors[p.dataIndex];
            }",
            paste(sprintf('\"%s\"', admins$party_color), collapse=",")
          ))
        )) |>
  e_y_axis(name = "%", axisLabel = list(formatter = "{value}%")) |>
  e_mark_line(data = list(yAxis = 0), lineStyle = list(color = "#6c757d"),
              symbol = "none") |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v === 'number' ? v.toFixed(1) + '%' : v; }")) |>
  e_title("S&P 500 total return over each administration",
          subtext = "First-full-month close to last-month close; nominal",
          left = "center") |>
  apply_econ_theme()
```

## S&P 500 timeline with administration bands {#timeline}

```{r sp-timeline}
#| echo: false
# Build markArea rectangles colored by party for each admin
bands <- admins |>
  transmute(
    start = format(window_first, "%Y-%m-%d"),
    end   = format(window_last,  "%Y-%m-%d"),
    color = party_color,
    label = label
  )

# Convert to the list-of-list structure markArea.data expects
mark_data <- lapply(seq_len(nrow(bands)), function(i) {
  list(
    list(
      xAxis = bands$start[i],
      itemStyle = list(color = paste0(bands$color[i], "33")),  # 20% opacity
      name = bands$label[i]
    ),
    list(xAxis = bands$end[i])
  )
})

monthly |>
  e_charts(date) |>
  e_line(sp500_level, name = "S&P 500", symbol = "none", smooth = TRUE,
         lineStyle = list(width = 2, color = "#1d3557")) |>
  e_mark_area(data = mark_data, silent = TRUE,
              label = list(show = TRUE, position = "insideTop",
                           color = "#495057", fontSize = 10)) |>
  e_tooltip(trigger = "axis") |>
  e_datazoom(x_index = 0) |>
  e_y_axis(name = "Index level", scale = TRUE) |>
  e_title("S&P 500 monthly level — colored by administration",
          left = "center") |>
  apply_econ_theme()
```

## Where to go next {#next}

- **[Economic Growth by Admin →](growth.qmd)** — GDP, unemployment, inflation
  tracked by administration.
- **[Markets by Admin →](markets.qmd)** — S&P 500, VIX, and volatility across
  administrations.
- **[Methodology & Caveats →](about.qmd)** — how administrations are assigned
  to months, limitations of per-administration comparisons, data sources.
