---
title: "Markets by Administration"
subtitle: "S&P 500, volatility, and returns per president"
---

```{r setup}
#| include: false
source("R/helpers.R")
source("R/presidential_helpers.R")

admins  <- load_admin_summary()
monthly <- load_monthly_admin() |> mutate(date = as.Date(date))

admins <- admins |>
  mutate(
    label       = admin_short_label(president, ongoing),
    party_color = party_color(party)
  )

color_array_js <- sprintf("[%s]",
  paste(sprintf('"%s"', admins$party_color), collapse = ","))

bar_color_js <- htmlwidgets::JS(sprintf(
  "function(p){ var c=%s; return c[p.dataIndex]; }", color_array_js))
```

::: {.callout-warning appearance="simple"}
**Market data on this site is a calibrated synthetic overlay** (FRED's market
series are licensed and cannot be redistributed). Directional comparisons
hold but exact levels are illustrative. See [Methodology](about.qmd) and the
[Data & Citations page](../economy/about.qmd) for details.
:::

## S&P 500 total return {#total}

```{r total-return}
#| echo: false
admins |>
  e_charts(label) |>
  e_bar(sp500_total_return, name = "Total return (%)",
        itemStyle = list(color = bar_color_js)) |>
  e_y_axis(name = "%", axisLabel = list(formatter = "{value}%")) |>
  e_mark_line(data = list(yAxis = 0), lineStyle = list(color = "#6c757d"),
              symbol = "none") |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? v.toFixed(1)+'%' : v; }")) |>
  e_title("Cumulative S&P 500 return over each administration's term",
          subtext = "First-full-month close to last-month close; nominal",
          left = "center") |>
  apply_econ_theme()
```

## S&P 500 annualized return {#annualized}

Normalizes total return by term length, so 4-year admins compare fairly to
8-year admins.

```{r annualized}
#| echo: false
admins |>
  e_charts(label) |>
  e_bar(sp500_annualized_return, name = "Annualized return (%)",
        itemStyle = list(color = bar_color_js)) |>
  e_y_axis(name = "%", axisLabel = list(formatter = "{value}%")) |>
  e_mark_line(data = list(yAxis = 0), lineStyle = list(color = "#6c757d"),
              symbol = "none") |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? v.toFixed(1)+'%' : v; }")) |>
  e_title("Annualized S&P 500 return per administration", left = "center") |>
  apply_econ_theme()
```

## Volatility — average VIX {#vix}

Higher VIX = more market uncertainty.

```{r vix}
#| echo: false
admins |>
  e_charts(label) |>
  e_bar(vix_avg, name = "Avg VIX",
        itemStyle = list(color = bar_color_js)) |>
  e_y_axis(name = "Index level") |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? v.toFixed(1) : v; }")) |>
  e_title("Average VIX per administration", left = "center") |>
  apply_econ_theme()
```

## Monthly S&P 500 level, administration-shaded {#timeline}

```{r sp-timeline}
#| echo: false
bands <- admins |>
  transmute(start = format(window_first, "%Y-%m-%d"),
            end   = format(window_last,  "%Y-%m-%d"),
            color = party_color, label)
mark_data <- lapply(seq_len(nrow(bands)), function(i) list(
  list(xAxis = bands$start[i],
       itemStyle = list(color = paste0(bands$color[i], "22")),
       name = bands$label[i]),
  list(xAxis = bands$end[i])
))

monthly |>
  e_charts(date) |>
  e_line(sp500_level, name = "S&P 500", symbol = "none", smooth = TRUE,
         lineStyle = list(width = 2, color = "#1d3557")) |>
  e_mark_area(data = mark_data, silent = TRUE,
              label = list(show = TRUE, position = "insideTop",
                           color = "#495057", fontSize = 10)) |>
  e_tooltip(trigger = "axis") |>
  e_datazoom(x_index = 0) |>
  e_y_axis(name = "Index level", scale = TRUE) |>
  e_title("S&P 500 monthly level — shaded by administration", left = "center") |>
  apply_econ_theme()
```
