---
title: "Economy vs Markets"
subtitle: "Do equities actually track economic fundamentals?"
---

```{r setup}
#| include: false
source("R/helpers.R")
register_econ_theme()

library(plotly)
library(crosstalk)

annual  <- load_annual()
monthly <- load_monthly()
sectors <- load_sectors()
```

A recurring question: how tightly is the stock market coupled to economic
growth? This page puts the two side-by-side, looks at correlations, and lets
you drill into S&P 500 sector performance by year.

## Annual GDP growth vs S&P 500 return {#dual}

Two very different pictures in the same window — equity returns are far more
volatile than GDP.

```{r dual-axis}
#| echo: false
annual |>
  mutate(label = ifelse(year >= 2025, paste0(year, "*"), as.character(year))) |>
  e_charts(label) |>
  e_bar(gdp_growth, name = "GDP growth (%)", itemStyle = list(color = "#2a6f97")) |>
  e_line(sp500_ret, name = "S&P 500 return (%)", y_index = 1,
         symbol = "circle", symbolSize = 7, smooth = FALSE,
         lineStyle = list(width = 2, color = "#bc4749")) |>
  e_mark_line(data = list(yAxis = 0), lineStyle = list(color = "#6c757d"),
              symbol = "none") |>
  e_y_axis(name = "GDP growth %", axisLabel = list(formatter = "{value}%")) |>
  e_y_axis(index = 1, name = "S&P return %", position = "right",
           axisLabel = list(formatter = "{value}%")) |>
  e_tooltip(trigger = "axis") |>
  e_title("GDP growth (bars) vs S&P 500 return (line)", left = "center") |>
  e_legend(top = 24) |>
  apply_econ_theme()
```

## How correlated are they, really? {#corr}

Each point is a year. Hover to see the year label. The 2020 COVID year and
the 2022 inflation shock pull the relationship around quite a bit.

```{r gdp-sp-scatter}
#| echo: false
library(plotly)

sd <- SharedData$new(annual, ~year, group = "yr")

plot_ly(sd,
        x = ~gdp_growth, y = ~sp500_ret, text = ~year,
        type = "scatter", mode = "markers+text",
        textposition = "top center",
        textfont = list(size = 10, color = "#6c757d"),
        marker = list(
          size = 12,
          color = ~sp500_ret,
          colorscale = list(c(0, "#bc4749"), c(0.5, "#f2c14e"), c(1, "#2f9e44")),
          line = list(color = "#1d3557", width = 1),
          showscale = FALSE
        ),
        hovertemplate = "<b>%{text}</b><br>GDP: %{x:.1f}%<br>S&P: %{y:.1f}%<extra></extra>") |>
  layout(
    xaxis = list(title = "Real GDP growth (%)", zeroline = TRUE),
    yaxis = list(title = "S&P 500 return (%)", zeroline = TRUE),
    shapes = list(
      list(type = "line", x0 = -4, x1 = 7, y0 = 0, y1 = 0,
           line = list(color = "#adb5bd", dash = "dot")),
      list(type = "line", x0 = 0, x1 = 0, y0 = -50, y1 = 50,
           line = list(color = "#adb5bd", dash = "dot"))
    ),
    margin = list(l = 60, r = 30, t = 30, b = 50)
  ) |>
  config(displayModeBar = FALSE)
```

```{r corr-values}
#| echo: false
h <- annual |> filter(year < 2026)
cors <- tibble::tibble(
  pair = c("GDP vs S&P 500", "GDP vs Dow", "GDP vs NASDAQ",
           "CPI vs S&P 500", "Unemployment vs S&P 500", "10Y yield vs S&P 500"),
  correlation = c(
    cor(h$gdp_growth, h$sp500_ret),
    cor(h$gdp_growth, h$dow_ret),
    cor(h$gdp_growth, h$nasdaq_ret),
    cor(h$cpi,        h$sp500_ret),
    cor(h$unemployment, h$sp500_ret),
    cor(h$ten_year,   h$sp500_ret)
  )
)

library(reactable)
reactable(
  cors,
  columns = list(
    pair = colDef(name = "Indicator pair (2001–2025)", minWidth = 240),
    correlation = colDef(
      name = "Pearson r",
      cell = function(v) {
        color <- if (v > 0) "#2f9e44" else "#bc4749"
        width <- paste0(round(abs(v) * 100), "%")
        htmltools::div(
          style = "display:flex; align-items:center; gap:10px;",
          htmltools::div(style = "width:50px; text-align:right; font-variant-numeric: tabular-nums; font-weight:600;",
                         sprintf("%+.2f", v)),
          htmltools::div(style = "flex:1; background:#e9ecef; height:8px; border-radius:4px;",
                         htmltools::div(style = sprintf("background:%s; height:8px; width:%s; border-radius:4px;",
                                                        color, width)))
        )
      },
      html = TRUE
    )
  ),
  striped = TRUE, compact = TRUE,
  theme = reactableTheme(borderColor = "#dee2e6")
)
```

## S&P 500 sector returns — drill by decade → year → sector {#sector-tree}

Treemap with native drill-down. Click a decade tile to zoom into its years;
click a year tile to see its sector-by-sector return breakdown. Sector tile
size is scaled by absolute return; color shows sign (green = positive,
red = negative).

```{r sector-treemap}
#| echo: false
library(purrr)

sec <- sectors |>
  mutate(decade = paste0(10 * (year %/% 10), "s"),
         abs_return = abs(return_pct))

# Build hierarchical treemap data: decade -> year -> sector
decades <- split(sec, sec$decade)

build_decade <- function(d) {
  years <- split(d, d$year)
  list(
    name = unique(d$decade),
    value = sum(d$abs_return),
    children = map(years, function(y) {
      list(
        name = as.character(unique(y$year)),
        value = sum(y$abs_return),
        children = map(seq_len(nrow(y)), function(i) {
          list(
            name = y$sector[i],
            value = y$abs_return[i],
            return_pct = y$return_pct[i],
            itemStyle = list(
              color = ifelse(y$return_pct[i] >= 0, "#2f9e44", "#bc4749")
            )
          )
        })
      )
    })
  )
}

tree_data <- map(decades, build_decade) |> unname()

e_charts() |>
  e_treemap(
    data = tree_data,
    name = "S&P 500 sector returns",
    roam = FALSE,
    leafDepth = 1,
    levels = list(
      list(itemStyle = list(borderColor = "#fff", borderWidth = 2, gapWidth = 2)),
      list(itemStyle = list(borderColor = "#f1f3f5", borderWidth = 1, gapWidth = 1),
           colorSaturation = c(0.35, 0.75)),
      list(colorSaturation = c(0.5, 0.9),
           itemStyle = list(borderColor = "#fff", borderWidth = 1, gapWidth = 1))
    ),
    breadcrumb = list(show = TRUE, top = "bottom")
  ) |>
  e_tooltip(formatter = htmlwidgets::JS("
    function(p){
      var r = p.data && p.data.return_pct;
      if (typeof r === 'number' && isFinite(r)) {
        var sign = r >= 0 ? '+' : '';
        return '<b>' + p.name + '</b><br/>Return: ' + sign + r.toFixed(1) + '%';
      }
      return '<b>' + p.name + '</b>';
    }
  ")) |>
  e_title("Sector returns — click to drill down", left = "center") |>
  apply_econ_theme()
```

## Cross-indicator heatmap {#heatmap}

Correlations across indicators, 2001–2025. Click a cell for the exact value.

```{r heatmap}
#| echo: false
cor_mat <- annual |>
  filter(year < 2026) |>
  select(gdp_growth, unemployment, cpi, fed_funds, ten_year,
         sp500_ret, dow_ret, nasdaq_ret, vix_avg) |>
  cor() |>
  round(2)

labels <- c("GDP growth", "Unemployment", "CPI", "Fed Funds", "10Y",
            "S&P 500", "Dow", "NASDAQ", "VIX")
dimnames(cor_mat) <- list(labels, labels)

cor_df <- as.data.frame(as.table(cor_mat)) |>
  setNames(c("x", "y", "value")) |>
  mutate(across(c(x, y), as.character))

cor_df |>
  e_charts(x) |>
  e_heatmap(y, value) |>
  e_visual_map(value, min = -1, max = 1,
               inRange = list(color = c("#bc4749", "#ffffff", "#2a6f97"))) |>
  e_tooltip(formatter = htmlwidgets::JS("
    function(p){
      var v = p.data && p.data[2];
      var r = (typeof v === 'number' && isFinite(v)) ? v.toFixed(2) : v;
      return p.data[0] + ' × ' + p.data[1] + '<br/>r = ' + r;
    }
  ")) |>
  e_title("Correlations across indicators (2001–2025)", left = "center") |>
  e_x_axis(axisLabel = list(rotate = 30)) |>
  apply_econ_theme()
```
