---
title: "Economic Output"
subtitle: "Annual GDP — Texas vs. national, with county-level output where available"
---

```{r setup}
#| include: false
source("R/helpers.R")
WT_DIR <- file.path(DATA_ROOT, "west-texas")

read_csv_safe <- function(p) {
  if (!file.exists(p) || file.size(p) == 0) return(tibble::tibble())
  suppressWarnings(readr::read_csv(p, show_col_types = FALSE))
}

gdp <- read_csv_safe(file.path(WT_DIR, "gdp_annual.csv"))
have_data <- nrow(gdp) > 0 && dplyr::n_distinct(gdp$geo) > 1
```

Annual gross domestic product from the Bureau of Economic Analysis. State
and national GDP data come from FRED (TXRGSP, GDPC1); county-level GDP
comes from BEA's CAGDP1 table.

::: {.callout-note}
## County GDP caveats

County GDP for very small counties may be suppressed by the BEA for
confidentiality. Where available, the data shows the total market value of
goods and services produced within the county — heavily influenced by oil and
gas extraction and ranching in this region.
:::

```{r gdp-growth-chart}
#| echo: false
#| eval: !expr have_data
#| fig-height: 7

library(echarts4r)

geo_order <- c("US", "TX", "sutton", "schleicher", "crockett", "kimble")
geo_colors <- c("#1d3557", "#2a6f97", "#e07a5f", "#6a4c93", "#2f9e44", "#f2c14e")
geo_names <- c(
  US         = "United States",
  TX         = "Texas",
  sutton     = "Sutton Co. (Sonora)",
  schleicher = "Schleicher Co. (Eldorado)",
  crockett   = "Crockett Co. (Ozona)",
  kimble     = "Kimble Co. (Junction)"
)

plot_data <- gdp |>
  dplyr::filter(!is.na(gdp_growth_pct)) |>
  dplyr::mutate(geo = factor(geo, levels = geo_order),
                geo_label = geo_names[as.character(geo)]) |>
  dplyr::filter(!is.na(geo))

yr_min <- min(plot_data$year, na.rm = TRUE)
yr_max <- max(plot_data$year, na.rm = TRUE)

plot_data |>
  dplyr::group_by(geo_label) |>
  e_charts(year) |>
  e_line(gdp_growth_pct, smooth = FALSE, symbol = "circle", symbolSize = 4) |>
  apply_econ_theme() |>
  e_title("Annual GDP Growth",
          subtext = "BEA CAGDP1 + FRED | Year-over-year % change") |>
  e_tooltip(trigger = "axis",
            formatter = htmlwidgets::JS("function(params) {
              var yr = params[0].axisValueLabel;
              var rows = params.filter(function(p) { return p.value[1] != null; }).map(function(p) {
                return '<tr><td>' + p.marker + ' ' + p.seriesName + '</td>' +
                '<td style=\"text-align:right;padding-left:12px\">' +
                p.value[1].toFixed(1) + '%</td></tr>';
              });
              return '<b>' + yr + '</b><table>' + rows.join('') + '</table>';
            }")) |>
  e_x_axis(type = "value", min = yr_min, max = yr_max,
           axisLabel = list(formatter = htmlwidgets::JS("function(v) { return Math.round(v); }"))) |>
  e_y_axis(name = "GDP Growth (%)",
           axisLabel = list(formatter = "{value}%")) |>
  e_legend(top = 62, type = "scroll") |>
  e_color(geo_colors)
```

```{r gdp-table}
#| echo: false
#| eval: !expr have_data

library(reactable)

gdp_wide <- gdp |>
  dplyr::mutate(geo_label = geo_names[geo]) |>
  dplyr::filter(!is.na(geo_label)) |>
  dplyr::select(year, geo_label, gdp) |>
  tidyr::pivot_wider(names_from = geo_label, values_from = gdp) |>
  dplyr::arrange(dplyr::desc(year))

reactable(gdp_wide,
          compact = TRUE,
          bordered = TRUE,
          striped = TRUE,
          defaultPageSize = 15,
          columns = lapply(names(gdp_wide)[-1], function(nm) {
            colDef(name = nm, format = colFormat(prefix = "$", separators = TRUE, digits = 0))
          }) |> setNames(names(gdp_wide)[-1]))
```

```{r no-data-msg}
#| echo: false
#| eval: !expr isFALSE(have_data)
htmltools::div(class = "callout callout-warning",
  htmltools::p("GDP data not yet available. The BEA API key must be ",
               "activated and the data pipeline run to populate this page. ",
               "Check the ", htmltools::a("Methodology", href = "about.html"),
               " page for details."))
```
