---
title: "Per-Capita Income"
subtitle: "Annual per-capita personal income — county vs. state vs. national"
---

```{r setup}
#| include: false
source("R/helpers.R")
WT_DIR <- file.path(DATA_ROOT, "west-texas")

read_csv_safe <- function(p) {
  if (!file.exists(p) || file.size(p) == 0) return(tibble::tibble())
  suppressWarnings(readr::read_csv(p, show_col_types = FALSE))
}

income <- read_csv_safe(file.path(WT_DIR, "income_annual.csv"))
have_data <- nrow(income) > 0 && dplyr::n_distinct(income$geo) > 1
```

Annual per-capita personal income from the Bureau of Economic Analysis (BEA)
Regional Economic Accounts. Income includes wages, proprietors' income,
dividends, interest, rent, and transfer receipts.

::: {.callout-note}
## BEA publication lag

County-level income data is published annually with a lag of approximately
6--9 months. The most recent year shown may be one to two years behind the
current calendar year.
:::

```{r income-chart}
#| echo: false
#| eval: !expr have_data
#| fig-height: 7

library(echarts4r)

geo_order <- c("US", "TX", "sutton", "schleicher", "crockett", "kimble")
geo_colors <- c("#1d3557", "#2a6f97", "#e07a5f", "#6a4c93", "#2f9e44", "#f2c14e")

plot_data <- income |>
  dplyr::mutate(geo = factor(geo, levels = geo_order)) |>
  dplyr::filter(!is.na(geo))

plot_data |>
  dplyr::group_by(geo_label) |>
  e_charts(year) |>
  e_line(per_capita_income, smooth = FALSE, symbol = "circle", symbolSize = 4) |>
  apply_econ_theme() |>
  e_title("Per-Capita Personal Income",
          subtext = "BEA CAINC1 | Annual, by county") |>
  e_tooltip(trigger = "axis",
            formatter = htmlwidgets::JS("function(params) {
              var yr = params[0].axisValueLabel;
              var rows = params.map(p =>
                '<tr><td>' + p.marker + ' ' + p.seriesName + '</td>' +
                '<td style=\"text-align:right;padding-left:12px\">$' +
                Math.round(p.value[1]).toLocaleString() + '</td></tr>');
              return '<b>' + yr + '</b><table>' + rows.join('') + '</table>';
            }")) |>
  e_y_axis(name = "Per-Capita Income ($)",
           axisLabel = list(formatter = htmlwidgets::JS(
             "function(v) { return '$' + (v/1000).toFixed(0) + 'k'; }"))) |>
  e_legend(top = 62, type = "scroll") |>
  e_color(geo_colors)
```

```{r income-table}
#| echo: false
#| eval: !expr have_data

library(reactable)

us_income <- income |> dplyr::filter(geo == "US")

income_wide <- income |>
  dplyr::select(year, geo_label, per_capita_income) |>
  tidyr::pivot_wider(names_from = geo_label, values_from = per_capita_income) |>
  dplyr::arrange(dplyr::desc(year))

reactable(income_wide,
          compact = TRUE,
          bordered = TRUE,
          striped = TRUE,
          defaultPageSize = 15,
          columns = lapply(names(income_wide)[-1], function(nm) {
            colDef(name = nm, format = colFormat(prefix = "$", separators = TRUE, digits = 0))
          }) |> setNames(names(income_wide)[-1]))
```

```{r no-data-msg}
#| echo: false
#| eval: !expr isFALSE(have_data)
htmltools::div(class = "callout callout-warning",
  htmltools::p("Income data not yet available. The BEA API key must be ",
               "activated and the data pipeline run to populate this page. ",
               "Check the ", htmltools::a("Methodology", href = "about.html"),
               " page for details."))
```
