---
title: "West Texas"
subtitle: "Regional economy — Sonora, Eldorado, Ozona, and Junction vs. Texas and the US"
---

```{r setup}
#| include: false
source("R/helpers.R")
WT_DIR <- file.path(DATA_ROOT, "west-texas")

read_csv_safe <- function(p) {
  if (!file.exists(p) || file.size(p) == 0) return(tibble::tibble())
  suppressWarnings(readr::read_csv(p, show_col_types = FALSE))
}

summ <- read_csv_safe(file.path(WT_DIR, "west_texas_summary.csv"))
```

A comparison of the West Texas regional economy to the state of Texas and
the United States, focusing on four small towns in the Edwards Plateau and
Permian Basin fringe:

- **Sonora** (Sutton County, pop. ~3,000)
- **Eldorado** (Schleicher County, pop. ~2,800)
- **Ozona** (Crockett County, pop. ~3,400)
- **Junction** (Kimble County, pop. ~4,400)

These communities share an economy shaped by ranching, oil and gas services,
and highway commerce. County-level data from the Bureau of Labor Statistics
and Bureau of Economic Analysis allows direct comparison to state and
national benchmarks.

```{r summary-cards}
#| echo: false
if (nrow(summ) > 0) {
  card_row(
    value_card("US Unemployment",
               fmt_pct(summ$us_ur),
               "Latest month, BLS",
               tone = "neutral"),
    value_card("Texas Unemployment",
               fmt_pct(summ$tx_ur),
               "Latest month, BLS",
               tone = "neutral"),
    value_card("West TX Region",
               fmt_pct(summ$regional_avg_ur),
               "Avg of 4 counties",
               tone = "neutral"),
    value_card("Regional Spread",
               fmt_signed_pct(round(summ$regional_avg_ur - summ$us_ur, 1)),
               "Region vs. US",
               tone = tone_by_sign(summ$regional_avg_ur - summ$us_ur,
                                   positive_tone = "negative",
                                   negative_tone = "positive"))
  )
} else {
  htmltools::div(class = "callout callout-warning",
    htmltools::p("West Texas dataset not yet populated. Run the data pipeline ",
                 "or wait for the next CI deploy."))
}
```

```{r income-cards}
#| echo: false
if (nrow(summ) > 0 && !is.na(summ$us_income)) {
  card_row(
    value_card("US Per-Capita Income",
               scales::dollar(summ$us_income, accuracy = 1),
               "Latest year, BEA",
               tone = "neutral"),
    value_card("Texas Per-Capita Income",
               scales::dollar(summ$tx_income, accuracy = 1),
               "Latest year, BEA",
               tone = "neutral"),
    value_card("West TX Region",
               scales::dollar(summ$regional_avg_income, accuracy = 1),
               "Avg of 4 counties",
               tone = "neutral")
  )
}
```

## Sections

- **[Unemployment →](unemployment.qmd)** — Monthly unemployment rates for
  each county vs. Texas and the US, from 2005 to present.
- **[Per-Capita Income →](income.qmd)** — Annual per-capita personal income
  by county, with comparisons to state and national averages.
- **[Economic Output →](gdp.qmd)** — Annual GDP for Texas and the US, with
  county-level output where available from the BEA.
- **[Methodology →](about.qmd)** — Data sources, FIPS codes, population
  context, and caveats about small-county volatility.

## Data sources

| Source | What | License | Refresh |
|---|---|---|---|
| [BLS LAUS](https://www.bls.gov/lau/) | Monthly county unemployment rates | Public domain (US gov) | Monthly |
| [BEA Regional](https://www.bea.gov/data/by-place-county-metro-local) | Annual county GDP and per-capita income | Public domain (US gov) | Annual (~6 mo lag) |
| [FRED](https://fred.stlouisfed.org/) | US unemployment (UNRATE), Texas unemployment (TXUR), Texas real GSP (TXRGSP) | Public domain | Daily / monthly |
