---
title: "Unemployment"
subtitle: "Monthly unemployment rates — US vs. Texas vs. West TX counties"
---

```{r setup}
#| include: false
source("R/helpers.R")
WT_DIR <- file.path(DATA_ROOT, "west-texas")

read_csv_safe <- function(p) {
  if (!file.exists(p) || file.size(p) == 0) return(tibble::tibble())
  suppressWarnings(readr::read_csv(p, show_col_types = FALSE))
}

unemp <- read_csv_safe(file.path(WT_DIR, "unemployment_monthly.csv"))
have_data <- nrow(unemp) > 0 && dplyr::n_distinct(unemp$geo) > 2
```

Monthly unemployment rates from the Bureau of Labor Statistics Local Area
Unemployment Statistics (LAUS) program. County-level data is available from
2005 onward for Sutton, Schleicher, Crockett, and Kimble counties.

::: {.callout-note}
## Small-county volatility

These counties have populations of 2,800--4,400. A single large employer
opening or closing can move the unemployment rate by several percentage
points. The national and state lines provide stable reference points; the
county lines show the real texture of the local labor market.
:::

```{r unemployment-chart}
#| echo: false
#| eval: !expr have_data
#| fig-height: 7

library(echarts4r)

# Order geos for consistent legend
geo_order <- c("US", "TX", "sutton", "schleicher", "crockett", "kimble")
geo_colors <- c(
  US         = "#1d3557",
  TX         = "#2a6f97",
  sutton     = "#e07a5f",
  schleicher = "#6a4c93",
  crockett   = "#2f9e44",
  kimble     = "#f2c14e"
)
geo_names <- c(
  US         = "United States",
  TX         = "Texas",
  sutton     = "Sutton Co. (Sonora)",
  schleicher = "Schleicher Co. (Eldorado)",
  crockett   = "Crockett Co. (Ozona)",
  kimble     = "Kimble Co. (Junction)"
)

plot_data <- unemp |>
  dplyr::mutate(geo = factor(geo, levels = geo_order),
                geo_label = geo_names[as.character(geo)]) |>
  dplyr::filter(!is.na(geo))

# Line widths: national/state thicker, counties thinner
line_widths <- c(US = 3, TX = 2.5, sutton = 1.5, schleicher = 1.5,
                 crockett = 1.5, kimble = 1.5)

chart <- plot_data |>
  dplyr::group_by(geo_label) |>
  e_charts(date) |>
  e_line(unemployment_rate, smooth = FALSE, symbol = "none") |>
  apply_econ_theme() |>
  e_title("Monthly Unemployment Rate",
          subtext = "BLS LAUS | County vs. state vs. national") |>
  e_tooltip(trigger = "axis",
            formatter = htmlwidgets::JS("function(params) {
              var d = params[0].axisValueLabel;
              var rows = params.map(p =>
                '<tr><td>' + p.marker + ' ' + p.seriesName + '</td>' +
                '<td style=\"text-align:right;padding-left:12px\">' +
                p.value[1].toFixed(1) + '%</td></tr>');
              return '<b>' + d + '</b><table>' + rows.join('') + '</table>';
            }")) |>
  e_x_axis(type = "time") |>
  e_y_axis(name = "Unemployment Rate (%)",
           axisLabel = list(formatter = "{value}%")) |>
  e_datazoom(type = "slider") |>
  e_legend(top = 62, type = "scroll") |>
  e_color(unname(geo_colors))

# Add recession bands
chart <- add_recession_bands(chart)
chart
```

```{r no-data-msg}
#| echo: false
#| eval: !expr isFALSE(have_data)
htmltools::div(class = "callout callout-warning",
  htmltools::p("Unemployment data not yet available. Run the data pipeline ",
               "to fetch BLS LAUS county data."))
```

```{r unemployment-table}
#| echo: false
#| eval: !expr have_data

library(reactable)

# Latest values by geo
latest <- unemp |>
  dplyr::group_by(geo) |>
  dplyr::filter(date == max(date)) |>
  dplyr::ungroup() |>
  dplyr::mutate(geo_label = geo_names[geo]) |>
  dplyr::arrange(match(geo, geo_order)) |>
  dplyr::select(Geography = geo_label, `Latest Rate (%)` = unemployment_rate,
                `As Of` = date)

reactable(latest,
          compact = TRUE,
          bordered = TRUE,
          striped = TRUE,
          columns = list(
            `Latest Rate (%)` = colDef(format = colFormat(digits = 1, suffix = "%")),
            `As Of` = colDef(format = colFormat(date = TRUE))
          ))
```
