---
title: "Economic Growth"
subtitle: "Real GDP — headline, quarterly detail, and components"
---

```{r setup}
#| include: false
source("R/helpers.R")

annual     <- load_annual()
quarterly  <- load_quarterly()
components <- load_gdp_components()

as_of_year <- max(annual$year)
```

Use the tabs below to drill from the annual headline into quarterly detail and
into the GDP expenditure components. The sunburst at the bottom lets you
drill into any single year's composition interactively (click a wedge to zoom
in; click the center to zoom out).

## Annual real GDP growth {#annual}

::: {.panel-tabset}

### Headline

```{r gdp-annual}
#| echo: false
annual |>
  mutate(
    color = ifelse(gdp_growth < 0, "#bc4749", "#2a6f97"),
    label = ifelse(prototype == 1, paste0(year, "*"), as.character(year))
  ) |>
  e_charts(label) |>
  e_bar(gdp_growth, name = "Real GDP growth (YoY)", itemStyle = list(
    color = htmlwidgets::JS("function(p){ return p.value[1] < 0 ? '#bc4749' : '#2a6f97'; }")
  )) |>
  e_mark_line(data = list(yAxis = mean(annual$gdp_growth[annual$prototype == 0], na.rm = TRUE)),
              lineStyle = list(type = "dashed", color = "#6c757d"),
              label = list(formatter = "Long-run avg", position = "end")) |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v === 'number' ? v.toFixed(2) + '%' : v; }")) |>
  e_title("Real GDP growth, YoY %", subtext = "* = current partial year", left = "center") |>
  e_y_axis(name = "%", axisLabel = list(formatter = "{value}%")) |>
  e_x_axis(axisLabel = list(rotate = 30, interval = 0)) |>
  e_datazoom(x_index = 0, start = 0, end = 100) |>
  apply_econ_theme()
```

### Quarterly detail

Every quarter since 1999 through the most recent available quarter.
Recession periods are shaded.

```{r gdp-quarterly}
#| echo: false
quarterly |>
  mutate(date = as.Date(date)) |>
  e_charts(date) |>
  e_line(gdp_growth, name = "Quarterly GDP growth (YoY proxy)",
         symbol = "circle", symbolSize = 6, smooth = TRUE,
         lineStyle = list(width = 2, color = "#2a6f97")) |>
  add_recession_bands(x_as_year = FALSE) |>
  e_mark_line(data = list(yAxis = 0),
              lineStyle = list(color = "#6c757d"), symbol = "none") |>
  e_tooltip(trigger = "axis") |>
  e_datazoom(x_index = 0) |>
  e_y_axis(name = "%", axisLabel = list(formatter = "{value}%")) |>
  e_title("Quarterly GDP growth", left = "center") |>
  apply_econ_theme()
```

### Components (share of GDP)

Stacked area showing the expenditure breakdown. Click a series name in the
legend to isolate.

```{r gdp-components}
#| echo: false
components |>
  pivot_longer(-year, names_to = "component", values_to = "share") |>
  mutate(component = recode(component,
    consumption = "Consumption (C)",
    investment  = "Investment (I)",
    government  = "Government (G)",
    net_exports = "Net exports (NX)")) |>
  group_by(component) |>
  e_charts(year) |>
  e_line(share, smooth = TRUE, symbol = "none",
         areaStyle = list(opacity = 0.35),
         stack = "total") |>
  e_tooltip(trigger = "axis") |>
  e_y_axis(name = "% of GDP", axisLabel = list(formatter = "{value}%")) |>
  e_title("GDP expenditure components, share of GDP", left = "center") |>
  e_legend(top = 56) |>
  apply_econ_theme()
```

:::

## Drill into any year — GDP composition sunburst {#sunburst}

Click a wedge to zoom in; click the center to zoom back out. Outer ring is the
four components; inner ring groups years by decade.

```{r gdp-sunburst}
#| echo: false
sb <- components |>
  mutate(decade = paste0(10 * (year %/% 10), "s")) |>
  pivot_longer(consumption:net_exports, names_to = "component", values_to = "share") |>
  mutate(
    component = recode(component,
      consumption = "Consumption",
      investment  = "Investment",
      government  = "Government",
      net_exports = "Net exports"),
    share = abs(share)
  )

hier <- lapply(split(sb, sb$decade), function(dec) {
  list(
    name = unique(dec$decade),
    children = lapply(split(dec, dec$year), function(yr) {
      list(
        name = as.character(unique(yr$year)),
        children = lapply(seq_len(nrow(yr)), function(i) {
          list(name = yr$component[i], value = yr$share[i])
        })
      )
    })
  )
})

e_charts() |>
  e_sunburst(
    data = hier,
    radius = c("10%", "95%"),
    itemStyle = list(borderColor = "#fff", borderWidth = 1),
    levels = list(
      list(),
      list(r0 = "10%", r = "35%",
           itemStyle = list(borderWidth = 2),
           label = list(rotate = "tangential", fontWeight = 600)),
      list(r0 = "35%", r = "60%",
           label = list(align = "right")),
      list(r0 = "60%", r = "90%",
           label = list(rotate = "radial"),
           itemStyle = list(borderWidth = 1))
    )
  ) |>
  e_title("GDP composition — drill by decade → year → component", left = "center") |>
  e_tooltip(formatter = htmlwidgets::JS("
    function(p){
      var v = p.value;
      if (typeof v === 'number' && isFinite(v)) {
        return p.name + ': ' + v.toFixed(1) + '%';
      }
      return p.name;
    }
  ")) |>
  apply_econ_theme()
```

## Recessions — how deep, how long? {#recessions}

```{r recession-depths}
#| echo: false
rec <- annual |>
  filter(recession == 1) |>
  select(year, gdp_growth, unemployment, sp500_ret)

library(reactable)
reactable(
  rec,
  columns = list(
    year         = colDef(name = "Year", maxWidth = 90),
    gdp_growth   = colDef(name = "GDP growth (%)",
                          format = colFormat(digits = 1, suffix = "%"),
                          style  = function(v) list(color = ifelse(v < 0, "#bc4749", "#2a6f97"),
                                                    fontWeight = 600)),
    unemployment = colDef(name = "Unemployment (%)",
                          format = colFormat(digits = 1, suffix = "%")),
    sp500_ret    = colDef(name = "S&P 500 return (%)",
                          format = colFormat(digits = 1, suffix = "%"),
                          style  = function(v) list(color = ifelse(v < 0, "#bc4749", "#2a6f97"),
                                                    fontWeight = 600))
  ),
  defaultSorted = list(year = "asc"),
  striped = TRUE, highlight = TRUE, compact = TRUE,
  theme = reactableTheme(borderColor = "#dee2e6")
)
```
