---
title: "US Economy 2026"
subtitle: "A visual analysis — 2026 in the context of the past 25 years"
---

```{r setup}
#| include: false
source("R/helpers.R")

annual  <- load_annual()
monthly <- load_monthly()

as_of_year <- max(annual$year)
current    <- annual |> filter(year == as_of_year)
history    <- annual |> filter(year < as_of_year & year >= as_of_year - 25)
```

An interactive drill-down analysis of the US economy placed against the
prior 25 years. Use the sidebar to move between panels, or click the links
at the bottom of this page.

::: {.callout-note appearance="simple"}
**Macro data** comes from FRED (public-domain US government series, refreshed
daily). **Market data** (S&P 500, Dow, NASDAQ, VIX, sector returns) is a
calibrated synthetic overlay — FRED's proprietary market licenses prohibit
redistribution of the raw series. The current (partial) year is flagged with
an asterisk. See [Data & Citations](about.qmd) for the full data dictionary
and source attribution.
:::

## 2026 Snapshot {#snapshot}

```{r snapshot-cards}
#| echo: false
cur <- current
hist <- history

gdp_pct <- percentile_vs_history(hist$gdp_growth, cur$gdp_growth)
ur_pct  <- percentile_vs_history(hist$unemployment, cur$unemployment)
cpi_pct <- percentile_vs_history(hist$cpi, cur$cpi)
sp_pct  <- percentile_vs_history(hist$sp500_ret, cur$sp500_ret)

card_row(
  value_card("Real GDP growth (YoY)",
             fmt_signed_pct(cur$gdp_growth),
             sprintf("%d\u1D57\u02B0 percentile vs prior 25 yrs", gdp_pct),
             tone = ifelse(cur$gdp_growth > 0, "positive", "negative")),
  value_card("Unemployment rate",
             fmt_pct(cur$unemployment),
             sprintf("%d\u1D57\u02B0 percentile (lower = better)", ur_pct),
             tone = ifelse(cur$unemployment < 5, "positive", "warn")),
  value_card("CPI inflation (YoY)",
             fmt_pct(cur$cpi),
             sprintf("%d\u1D57\u02B0 percentile", cpi_pct),
             tone = ifelse(cur$cpi < 3, "positive", "warn")),
  value_card("Fed Funds target",
             fmt_pct(cur$fed_funds, 2),
             "Upper bound, period avg",
             tone = "neutral"),
  value_card("10Y Treasury",
             fmt_pct(cur$ten_year, 2),
             "Period avg",
             tone = "neutral"),
  value_card("S&P 500 (YTD)",
             fmt_signed_pct(cur$sp500_ret),
             sprintf("%d\u1D57\u02B0 percentile", sp_pct),
             tone = ifelse(cur$sp500_ret > 0, "positive", "negative"))
)
```

## Where does the current year sit in the 25-year range? {#range}

Each bar shows the prior-25-year range for a given indicator, with the
current (partial) year marked.

```{r range-plot}
#| echo: false
indicators <- tibble::tribble(
  ~name,                ~col,            ~unit,
  "GDP growth",         "gdp_growth",    "%",
  "Unemployment",       "unemployment",  "%",
  "CPI inflation",      "cpi",           "%",
  "Fed Funds",          "fed_funds",     "%",
  "10Y Treasury",       "ten_year",      "%",
  "S&P 500 return",     "sp500_ret",     "%",
  "VIX (avg)",          "vix_avg",       ""
)

range_df <- purrr::pmap_dfr(indicators, function(name, col, unit) {
  h <- history[[col]]
  tibble(
    indicator = name,
    min_25 = min(h),
    max_25 = max(h),
    median_25 = median(h),
    current = current[[col]]
  )
})

library(plotly)

plot_ly(range_df) |>
  add_trace(
    x = ~indicator,
    y = ~(max_25 - min_25),
    base = ~min_25,
    type = "bar",
    name = "Prior 25-yr range",
    marker = list(color = "rgba(42,111,151,0.28)",
                  line = list(color = "#2a6f97", width = 1)),
    hovertemplate = "%{x}<br>Range: %{base:.2f}% – %{customdata:.2f}%<extra></extra>",
    customdata = ~max_25,
    width = 0.35
  ) |>
  add_trace(
    x = ~indicator, y = ~median_25,
    type = "scatter", mode = "markers",
    name = "25-yr median",
    marker = list(symbol = "diamond", size = 11, color = "#1d3557"),
    hovertemplate = "%{x}<br>Median: %{y:.2f}%<extra></extra>"
  ) |>
  add_trace(
    x = ~indicator, y = ~current,
    type = "scatter", mode = "markers",
    name = paste0(as_of_year, "*"),
    marker = list(symbol = "circle", size = 14, color = "#bc4749",
                  line = list(color = "#7a1e1f", width = 1)),
    hovertemplate = paste0("%{x}<br>", as_of_year, ": %{y:.2f}%<extra></extra>")
  ) |>
  layout(
    title = list(text = paste0(as_of_year, " values vs prior 25-year range"),
                 x = 0.5, xanchor = "center"),
    xaxis = list(title = "", tickangle = -25),
    yaxis = list(title = "%", zeroline = TRUE, zerolinecolor = "#adb5bd"),
    legend = list(orientation = "h", y = 1.1, x = 0.5, xanchor = "center"),
    margin = list(l = 50, r = 30, t = 60, b = 80),
    barmode = "overlay"
  ) |>
  config(displayModeBar = FALSE)
```

## S&P 500 over the 25-year window {#sp500}

Recession periods are shaded. Use the scrollbar below the chart to zoom.

```{r sp500-monthly}
#| echo: false
m <- monthly |> mutate(date = as.Date(date))

m |>
  e_charts(date) |>
  e_line(sp500_level, name = "S&P 500", symbol = "none", smooth = TRUE,
         lineStyle = list(width = 2, color = "#2a6f97"),
         areaStyle = list(color = "rgba(42,111,151,0.10)")) |>
  add_recession_bands(x_as_year = FALSE) |>
  e_tooltip(trigger = "axis") |>
  e_datazoom(x_index = 0) |>
  e_y_axis(name = "Index level", scale = TRUE) |>
  e_title("S&P 500 — monthly level", left = "center") |>
  apply_econ_theme()
```

## Where to go next {#next}

- **[Economic Growth →](growth.qmd)** — GDP growth, quarterly detail, and GDP components (consumption / investment / government / net exports).
- **[Indicators →](indicators.qmd)** — Unemployment, inflation, rates, VIX — filterable table with inline sparklines; click a row for the full time series.
- **[Economy vs Markets →](markets.qmd)** — Do markets actually track economic growth? Correlation scatter, dual-axis view, and sector treemap.
- **[Data & Citations →](about.qmd)** — Data dictionary, source attribution, and licensing.
