---
title: "Economic Indicators"
subtitle: "Unemployment, inflation, rates, VIX — 1999 to present"
---

```{r setup}
#| include: false
source("R/helpers.R")

annual  <- load_annual()
monthly <- load_monthly()

as_of_year <- max(annual$year)
```

Each indicator is shown two ways:

- **Summary table** — current-year value, historical median, historical percentile, inline sparkline. Click the chevron on a row to see the full annual chart with recession shading.
- **Comparative monthly view** — all indicators on one normalized chart; toggle series in the legend to compare.

## Summary table {#table}

```{r indicator-table}
#| echo: false
library(reactable)
library(htmltools)
library(sparkline)

history <- annual |> filter(year < as_of_year)
current <- annual |> filter(year == as_of_year)

indicators <- tibble::tribble(
  ~col,           ~label,                  ~unit,
  "gdp_growth",   "Real GDP growth (YoY)", "%",
  "unemployment", "Unemployment rate",     "%",
  "cpi",          "CPI inflation (YoY)",   "%",
  "fed_funds",    "Fed Funds rate",        "%",
  "ten_year",     "10Y Treasury yield",    "%",
  "vix_avg",      "VIX (annual avg)",      ""
) |>
  mutate(
    current_val = purrr::map_dbl(col, ~ current[[.x]]),
    median_25   = purrr::map_dbl(col, ~ median(history[[.x]])),
    min_25      = purrr::map_dbl(col, ~ min(history[[.x]])),
    max_25      = purrr::map_dbl(col, ~ max(history[[.x]])),
    pct_25      = purrr::map2_dbl(col, current_val,
                                  ~ percentile_vs_history(history[[.x]], .y)),
    sparkvals   = purrr::map(col, ~ c(history[[.x]], current[[.x]]))
  )

indicators_tbl <- indicators |>
  select(label, current_val, median_25, min_25, max_25, pct_25, sparkvals, col)

reactable(
  indicators_tbl,
  defaultSorted = list(pct_25 = "desc"),
  striped = TRUE, highlight = TRUE, compact = TRUE,
  theme = reactableTheme(borderColor = "#dee2e6"),
  columns = list(
    label = colDef(name = "Indicator", minWidth = 180),
    current_val = colDef(name = paste0(as_of_year, "*"),
                         format = colFormat(digits = 2, suffix = "%"),
                         style = list(fontWeight = 600)),
    median_25 = colDef(name = "Median",
                       format = colFormat(digits = 2, suffix = "%")),
    min_25 = colDef(name = "Min",
                    format = colFormat(digits = 2, suffix = "%")),
    max_25 = colDef(name = "Max",
                    format = colFormat(digits = 2, suffix = "%")),
    pct_25 = colDef(
      name = "Percentile",
      cell = function(v) {
        pct <- paste0(v, "%")
        bar_w <- paste0(v, "%")
        bar <- div(style = sprintf(
          "background:#2a6f97; height:8px; width:%s; border-radius:4px;", bar_w))
        tagList(
          div(style = "display:flex; align-items:center; gap:8px;",
              div(style = "flex:1; background:#e9ecef; height:8px; border-radius:4px;", bar),
              div(style = "width:38px; text-align:right; font-variant-numeric: tabular-nums;", pct))
        )
      },
      html = TRUE
    ),
    sparkvals = colDef(
      name = "Trend",
      cell = function(values) {
        sparkline(values, type = "line", width = 150, height = 30,
                  lineColor = "#2a6f97", fillColor = "#d0e2ef",
                  spotColor = "#bc4749", minSpotColor = "#6c757d", maxSpotColor = "#2f9e44")
      },
      html = TRUE,
      width = 180
    ),
    col = colDef(show = FALSE)
  ),
  details = function(index) {
    row <- indicators_tbl[index, ]
    col_name <- row$col
    series <- annual |> transmute(year, value = .data[[col_name]], prototype)
    chart <- series |>
      e_charts(year) |>
      e_line(value, name = row$label, symbol = "circle", symbolSize = 7,
             smooth = TRUE,
             lineStyle = list(width = 2, color = "#2a6f97"),
             areaStyle = list(color = "rgba(42,111,151,0.12)")) |>
      add_recession_bands(x_as_year = TRUE) |>
      e_mark_line(data = list(yAxis = median(history[[col_name]])),
                  lineStyle = list(type = "dashed", color = "#1d3557"),
                  label = list(formatter = "Historical median", position = "end")) |>
      e_tooltip(trigger = "axis") |>
      e_y_axis(axisLabel = list(formatter = "{value}%")) |>
      e_title(paste0(row$label, " — annual"), left = "center") |>
      apply_econ_theme()
    htmltools::div(style = "padding: 8px 6px 14px;", chart)
  }
)
```

<small>\* = current partial year; see [Data & Citations](about.qmd) for full provenance.</small>

## Comparative monthly view {#compare}

Monthly series from 1999 to present. Start with unemployment selected; use
the legend to toggle others on/off.

```{r indicator-compare}
#| echo: false
m <- monthly |> mutate(date = as.Date(date))

m |>
  e_charts(date) |>
  e_line(unemployment, name = "Unemployment (%)", symbol = "none", smooth = TRUE) |>
  e_line(cpi,          name = "CPI YoY (%)",      symbol = "none", smooth = TRUE) |>
  e_line(fed_funds,    name = "Fed Funds (%)",    symbol = "none", smooth = TRUE) |>
  e_line(ten_year,     name = "10Y Treasury (%)", symbol = "none", smooth = TRUE) |>
  e_line(vix,          name = "VIX",              symbol = "none", smooth = TRUE, y_index = 1) |>
  add_recession_bands(x_as_year = FALSE) |>
  e_y_axis(name = "%", axisLabel = list(formatter = "{value}%")) |>
  e_y_axis(index = 1, name = "VIX", position = "right", axisLabel = list(formatter = "{value}")) |>
  e_tooltip(trigger = "axis") |>
  e_datazoom(x_index = 0) |>
  e_legend(selected = list(
    "Unemployment (%)" = TRUE,
    "CPI YoY (%)"      = TRUE,
    "Fed Funds (%)"    = FALSE,
    "10Y Treasury (%)" = FALSE,
    "VIX"              = FALSE
  ), top = 24) |>
  e_title("Monthly indicators — recession bands shaded", left = "center") |>
  apply_econ_theme()
```

## Rate environment in context {#rates}

Fed Funds vs 10Y Treasury. When the curve inverts (Fed Funds > 10Y), a
recession has historically followed within ~18 months.

```{r rates-spread}
#| echo: false
m |>
  mutate(spread = ten_year - fed_funds) |>
  e_charts(date) |>
  e_line(ten_year,  name = "10Y Treasury (%)", symbol = "none", smooth = TRUE) |>
  e_line(fed_funds, name = "Fed Funds (%)",    symbol = "none", smooth = TRUE) |>
  e_line(spread,    name = "Term spread (10Y − FF)", symbol = "none", smooth = TRUE,
         lineStyle = list(type = "dashed"), y_index = 1) |>
  e_mark_line(data = list(yAxis = 0), lineStyle = list(color = "#bc4749"),
              symbol = "none") |>
  add_recession_bands(x_as_year = FALSE) |>
  e_y_axis(name = "Yield (%)", axisLabel = list(formatter = "{value}%")) |>
  e_y_axis(index = 1, name = "Spread (pp)", position = "right") |>
  e_tooltip(trigger = "axis") |>
  e_datazoom(x_index = 0) |>
  e_title("Rate environment: yields and term spread", left = "center") |>
  apply_econ_theme()
```
