---
title: "Unemployment — 1999 to Present"
subtitle: "Monthly rate with administration bands and rolling averages"
---

```{r setup}
#| include: false
source("R/helpers.R")
source("R/presidential_helpers.R")

monthly <- load_monthly() |>
  dplyr::select(date, unemployment) |>
  dplyr::mutate(date = as.Date(date)) |>
  dplyr::arrange(date) |>
  dplyr::filter(!is.na(unemployment))

admins <- load_administrations() |>
  dplyr::mutate(
    start_date    = as.Date(start_date),
    effective_end = as.Date(effective_end)
  )

# ---- Rolling averages --------------------------------------------------
# We use simple trailing means on the monthly series. 1y = 12 months,
# 5y = 60, 10y = 120. `zoo::rollmeanr` aligns means right (each point
# is the mean of the N preceding months including itself).
roll <- function(x, k) {
  out <- rep(NA_real_, length(x))
  for (i in seq_along(x)) {
    if (i >= k) out[i] <- mean(x[(i - k + 1):i], na.rm = TRUE)
  }
  out
}

# Pandemic window to exclude from the "adjusted" average: March 2020 to
# December 2021 (NBER recession ended April 2020 but labor-market
# distortion persisted).
pandemic_mask <- monthly$date >= as.Date("2020-03-01") &
                 monthly$date <= as.Date("2021-12-31")

monthly <- monthly |>
  dplyr::mutate(
    avg_1y  = roll(unemployment, 12),
    avg_5y  = roll(unemployment, 60),
    avg_10y = roll(unemployment, 120),
    # Pandemic-excluded long-run average: running mean of non-pandemic months
    adj_val = ifelse(pandemic_mask, NA_real_, unemployment)
  )
# Compute adjusted 5y average: rolling mean of non-pandemic values only
monthly$avg_5y_adj <- NA_real_
for (i in seq_len(nrow(monthly))) {
  window_start <- max(1, i - 59)
  window <- monthly$adj_val[window_start:i]
  if (sum(!is.na(window)) >= 24) {
    monthly$avg_5y_adj[i] <- mean(window, na.rm = TRUE)
  }
}
```

The heavy black line is **actual unemployment** (BLS civilian unemployment
rate, monthly). Dashed colored lines are **trailing rolling averages**:
1-year, 5-year, and 10-year. The dashed grey line is a **pandemic-adjusted**
5-year average that excludes March 2020 – December 2021, the window where
the labor market was severely distorted by COVID-19 and its aftermath.

Colored vertical bands show each presidential administration's term.

```{r plot}
#| echo: false
suppressPackageStartupMessages(library(echarts4r))

# Per-admin shaded bands
bands <- admins |>
  dplyr::mutate(
    color = party_color(party),
    label = admin_short_label(president, ongoing),
    clip_start = pmax(start_date,    min(monthly$date)),
    clip_end   = pmin(effective_end, max(monthly$date))
  ) |>
  dplyr::filter(clip_end > clip_start)

mark_bands <- lapply(seq_len(nrow(bands)), function(i) list(
  list(xAxis = format(bands$clip_start[i], "%Y-%m-%d"),
       itemStyle = list(color = paste0(bands$color[i], "22")),
       name = bands$label[i]),
  list(xAxis = format(bands$clip_end[i], "%Y-%m-%d"))
))

# Pandemic window — appended to the same list so a single e_mark_area call
# handles everything (multiple calls to e_mark_area on the same series fight
# each other in echarts).
mark_bands <- c(mark_bands, list(list(
  list(xAxis = "2020-03-01",
       itemStyle = list(color = "rgba(29,53,87,0.12)"),
       name = "Pandemic (excluded)"),
  list(xAxis = "2021-12-31")
)))

monthly |>
  e_charts(date) |>
  e_line(unemployment, name = "Unemployment (actual)",
         symbol = "none", smooth = FALSE,
         lineStyle = list(width = 2.4, color = "#1d1d1d")) |>
  e_line(avg_1y,  name = "1-yr rolling avg",
         symbol = "none", smooth = TRUE,
         lineStyle = list(width = 1.6, color = "#2f9e44", type = "dashed")) |>
  e_line(avg_5y,  name = "5-yr rolling avg",
         symbol = "none", smooth = TRUE,
         lineStyle = list(width = 1.6, color = "#e07a5f", type = "dashed")) |>
  e_line(avg_10y, name = "10-yr rolling avg",
         symbol = "none", smooth = TRUE,
         lineStyle = list(width = 1.6, color = "#6a4c93", type = "dashed")) |>
  e_line(avg_5y_adj, name = "5-yr adj (excl. pandemic)",
         symbol = "none", smooth = TRUE,
         lineStyle = list(width = 1.6, color = "#6c757d", type = "dotted")) |>
  e_mark_area(data = mark_bands, silent = TRUE,
              label = list(show = TRUE, position = "insideTop",
                           color = "#495057", fontSize = 10)) |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS(
              "function(v){ return typeof v==='number' ? v.toFixed(2)+'%' : v; }")) |>
  e_datazoom(x_index = 0) |>
  e_y_axis(name = "%", axisLabel = list(formatter = "{value}%")) |>
  e_legend(top = 56) |>
  e_title("US unemployment rate — actual vs trailing averages",
          subtext = "Administration bands + pandemic exclusion window",
          left = "center") |>
  apply_econ_theme()
```

## Reading the chart

- **Heavy black** — the raw monthly rate. Every movement you see is a BLS
  release.
- **1-year (green dashed)** — smooths away monthly noise but still follows
  cyclical turns. Useful for "where are we *really* right now?"
- **5-year (orange dashed)** — removes business-cycle noise. A rising 5-yr
  average signals structural labor-market deterioration; a falling one
  signals structural tightness.
- **10-year (purple dashed)** — secular trend; moves slowly. Currently
  dominated by the post-GFC recovery long plateau and the 2020 spike.
- **5-year adjusted (grey dotted)** — same 5-year window but excludes
  March 2020 – December 2021 observations. Shows what "normal" labor-market
  conditions looked like minus the pandemic shock. Useful when judging
  whether the current labor market is "tight" versus pre-pandemic norms
  rather than pandemic-distorted norms.

## Why the pandemic exclusion matters

March 2020 – April 2020 saw unemployment spike from 3.5% to 14.7% in a
single month — a once-in-a-century labor-market shock driven by public-
health policy, not underlying economic conditions. Including those months
in a trailing 5-year average keeps them in every window until ~March 2025,
distorting any "how does the current rate compare to recent history"
judgment throughout the Biden term and most of Trump's second term.

The adjusted line answers: **"Compared to the pre-pandemic trend, how
does the current labor market look?"**

## Data source

BLS Civilian Unemployment Rate (FRED series `UNRATE`), monthly. Loaded
from `data/economy/monthly.csv` which is refreshed daily from FRED.
Administration boundaries from the presidential-economies analysis's
`administrations.csv`.
