---
title: "Political Approval"
subtitle: "Gallup per-administration averages, 1999 to present"
---

```{r setup}
#| include: false
source("R/helpers.R")
source("R/presidential_helpers.R")

sent <- load_admin_sentiment() |> filter(!is.na(party))
gallup <- load_gallup_approval()

sent <- sent |>
  mutate(label       = admin_short_label(president, ongoing),
         party_color = party_color(party))

color_array_js <- sprintf("[%s]",
  paste(sprintf('"%s"', sent$party_color), collapse = ","))
bar_color_js <- htmlwidgets::JS(sprintf(
  "function(p){ var c=%s; return c[p.dataIndex]; }", color_array_js))

baseline <- round(mean(gallup$avg_approval, na.rm = TRUE), 1)
```

Gallup's *job approval* question is the longest continuous polling series of
its kind. The numbers below are **term-average approval ratings** — not
snapshot polls. They tell you what percentage of Americans, on average
across each president's entire term, answered "approve" when asked whether
they approved of the job the president was doing.

::: {.callout-note appearance="simple"}
Gallup approval is not available from a free API; the values in
`gallup_approval.csv` are **hand-curated** from Gallup's published summaries
and updated on demand, not daily. The *economic-sentiment* portion of this
site refreshes daily from FRED; this section does not.
:::

## Per-administration average {#avg}

Horizontal reference line: the average of completed admins' averages
(`r baseline`%).

```{r avg-bar}
#| echo: false
sent |>
  filter(!is.na(gallup_avg)) |>
  e_charts(label) |>
  e_bar(gallup_avg, name = "Gallup approval avg (%)",
        itemStyle = list(color = bar_color_js)) |>
  e_mark_line(data = list(yAxis = baseline),
              lineStyle = list(type = "dashed", color = "#6c757d"),
              label = list(formatter = sprintf("Baseline (%.1f%%)", baseline),
                           position = "end")) |>
  e_y_axis(name = "%", axisLabel = list(formatter = "{value}%"), min = 30, max = 70) |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? v.toFixed(1)+'%' : v; }")) |>
  e_title("Gallup approval: term average per administration",
          left = "center") |>
  apply_econ_theme()
```

## Range: min to max during each term {#range}

Presidents often peak early (honeymoon) or after major rallying events (9/11),
and decline late. The bars below show the full min–max range each president
traversed, with a dot marking their term-average.

```{r range-plot}
#| echo: false
library(plotly)

rng <- sent |>
  filter(!is.na(gallup_avg)) |>
  select(label, party_color, gallup_min, gallup_avg, gallup_max, gallup_last)

plot_ly(rng) |>
  add_trace(
    x = ~label, y = ~(gallup_max - gallup_min), base = ~gallup_min,
    type = "bar", name = "Min – Max range",
    marker = list(color = ~sapply(party_color, function(c) paste0(c, "66")),
                  line = list(color = ~party_color, width = 1)),
    width = 0.4,
    hovertemplate = "%{x}<br>Min: %{base:.0f}%<br>Max: %{customdata:.0f}%<extra></extra>",
    customdata = ~gallup_max
  ) |>
  add_trace(
    x = ~label, y = ~gallup_avg, type = "scatter", mode = "markers",
    name = "Term average",
    marker = list(symbol = "diamond", size = 11, color = "#1d3557"),
    hovertemplate = "%{x}<br>Avg: %{y:.1f}%<extra></extra>"
  ) |>
  add_trace(
    x = ~label, y = ~gallup_last, type = "scatter", mode = "markers",
    name = "Exit approval",
    marker = list(symbol = "circle", size = 11, color = "#bc4749"),
    hovertemplate = "%{x}<br>Exit: %{y:.0f}%<extra></extra>"
  ) |>
  layout(
    xaxis = list(title = ""),
    yaxis = list(title = "Approval (%)", range = c(20, 95)),
    legend = list(orientation = "h", y = 1.1, x = 0.5, xanchor = "center"),
    margin = list(l = 50, r = 30, t = 30, b = 50),
    barmode = "overlay"
  ) |>
  config(displayModeBar = FALSE)
```

## Detailed Gallup table {#table}

```{r detail-table}
#| echo: false
library(reactable); library(htmltools)

party_badge <- function(party) {
  col <- party_color(party)
  tags$span(
    style = sprintf(
      "display:inline-block; padding:2px 8px; border-radius:3px; font-size:0.72rem;
       font-weight:600; color:#fff; background:%s;", col),
    toupper(party)
  )
}

gallup |>
  transmute(
    president = paste0(president, ifelse(ongoing, " *", "")),
    party,
    avg_approval, min_approval, max_approval, last_approval, notes
  ) |>
  reactable(
    striped = TRUE, highlight = TRUE, compact = TRUE, bordered = FALSE,
    theme = reactableTheme(borderColor = "#dee2e6"),
    columns = list(
      president = colDef(name = "President", minWidth = 180),
      party = colDef(name = "Party", minWidth = 110,
                     cell = function(v) party_badge(v), html = TRUE),
      avg_approval = colDef(name = "Avg (%)",
                            format = colFormat(digits = 1, suffix = "%"),
                            style = list(fontWeight = 600)),
      min_approval = colDef(name = "Min (%)",
                            format = colFormat(suffix = "%")),
      max_approval = colDef(name = "Max (%)",
                            format = colFormat(suffix = "%")),
      last_approval = colDef(name = "Exit (%)",
                             format = colFormat(suffix = "%")),
      notes = colDef(name = "Notes", minWidth = 320)
    )
  )
```

## Sources

- **Gallup News:** *Presidential Job Approval Center.*
  <https://news.gallup.com/interactives/185273/presidential-job-approval-center.aspx>
- **The American Presidency Project (UCSB):** *Presidential Job Approval Tables.*
  <https://www.presidency.ucsb.edu/statistics/data/presidential-job-approval>

Aggregate averages are facts in the *Feist v. Rural* sense (not copyrightable),
but survey methodology and phrasing vary across polling firms. These figures
use Gallup's consistent methodology throughout.
