---
title: "Economic Sentiment"
subtitle: "U. Michigan Consumer Sentiment Index, 1999 to present"
---

```{r setup}
#| include: false
source("R/helpers.R")
source("R/presidential_helpers.R")

sent    <- load_admin_sentiment() |> filter(!is.na(party))
monthly <- load_umcsent_monthly()
events  <- load_world_events()

sent <- sent |>
  mutate(label       = admin_short_label(president, ongoing),
         party_color = party_color(party))

color_array_js <- sprintf("[%s]",
  paste(sprintf('"%s"', sent$party_color), collapse = ","))
bar_color_js <- htmlwidgets::JS(sprintf(
  "function(p){ var c=%s; return c[p.dataIndex]; }", color_array_js))

baseline <- round(mean(monthly$umcsent, na.rm = TRUE), 1)
```

The **University of Michigan Index of Consumer Sentiment (UMCSENT)** has been
published monthly since 1952. It's a survey-based index — households answer
five questions about their own financial situation, expected business
conditions, and attitudes toward major purchases. The output is a single
number; higher = more optimistic. It's the most widely-cited measure of
"what the public feels about the economy."

::: {.callout-note appearance="simple"}
UMCSENT is a measure of **economic mood**, not presidential approval. It
correlates with, but is not the same as, how the public feels about the
president. A president can be unpopular while the economy feels good
(Trump I, 2018-19) or popular while it feels bad (Bush 43 post-9/11, 2001).
:::

## UMCSENT timeline, administration-shaded {#timeline}

World events are marked on the time axis. Click the data-zoom slider to focus
on a specific period.

```{r timeline}
#| echo: false
# Admin-colored bands
admins_ref <- load_administrations() |>
  mutate(start_date = as.Date(start_date),
         effective_end = as.Date(effective_end))

bands <- admins_ref |>
  mutate(
    color = party_color(party),
    label = admin_short_label(president, ongoing),
    # clip to UMCSENT monthly window
    clip_start = pmax(start_date, min(monthly$date)),
    clip_end   = pmin(effective_end, max(monthly$date))
  ) |>
  filter(clip_end > clip_start)

mark_bands <- lapply(seq_len(nrow(bands)), function(i) list(
  list(xAxis = format(bands$clip_start[i], "%Y-%m-%d"),
       itemStyle = list(color = paste0(bands$color[i], "22")),
       name = bands$label[i]),
  list(xAxis = format(bands$clip_end[i],   "%Y-%m-%d"))
))

# Event markers as vertical lines
event_lines <- lapply(seq_len(nrow(events)), function(i) list(
  xAxis = format(events$date[i], "%Y-%m-%d"),
  label = list(
    formatter = events$event[i],
    position  = "insideEndTop",
    rotate    = 90,
    fontSize  = 9,
    color     = "#6c757d"
  ),
  lineStyle = list(
    color = "#495057", type = "dashed", width = 1, opacity = 0.6
  )
))

monthly |>
  e_charts(date) |>
  e_line(umcsent, name = "UMCSENT", symbol = "none", smooth = TRUE,
         lineStyle = list(width = 2, color = "#1d3557")) |>
  e_mark_area(data = mark_bands, silent = TRUE,
              label = list(show = TRUE, position = "insideTop",
                           color = "#495057", fontSize = 10)) |>
  e_mark_line(data = event_lines, symbol = "none", silent = TRUE) |>
  e_mark_line(data = list(yAxis = baseline),
              lineStyle = list(type = "dotted", color = "#bc4749"),
              label = list(formatter = sprintf("Baseline %.0f", baseline),
                           position = "end")) |>
  e_tooltip(trigger = "axis") |>
  e_datazoom(x_index = 0) |>
  e_y_axis(name = "Index", scale = TRUE) |>
  e_title("UMCSENT monthly — shaded by administration, marked world events",
          left = "center") |>
  apply_econ_theme()
```

## Per-administration UMCSENT average {#avg}

Horizontal reference line is the long-run baseline across the window
(`r baseline`). Trump I's elevated average reflects the pre-COVID expansion;
the Biden and Trump II averages sit well below baseline — inflation-era mood.

```{r admin-bars}
#| echo: false
sent |>
  filter(!is.na(umcsent_avg)) |>
  e_charts(label) |>
  e_bar(umcsent_avg, name = "UMCSENT avg",
        itemStyle = list(color = bar_color_js)) |>
  e_mark_line(data = list(yAxis = baseline),
              lineStyle = list(type = "dashed", color = "#6c757d"),
              label = list(formatter = sprintf("Baseline %.1f", baseline),
                           position = "end")) |>
  e_y_axis(name = "Index", min = 40, max = 120) |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? v.toFixed(1) : v; }")) |>
  e_title("UMCSENT: term-average sentiment per administration",
          left = "center") |>
  apply_econ_theme()
```

## Vs baseline {#vs-baseline}

```{r vs-baseline}
#| echo: false
sent |>
  filter(!is.na(umcsent_vs_baseline)) |>
  e_charts(label) |>
  e_bar(umcsent_vs_baseline, name = "vs baseline",
        itemStyle = list(color = bar_color_js)) |>
  e_mark_line(data = list(yAxis = 0), lineStyle = list(color = "#6c757d"),
              symbol = "none") |>
  e_y_axis(name = "Index points") |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? (v > 0 ? '+' : '')+v.toFixed(1) : v; }")) |>
  e_title("UMCSENT average minus long-run baseline",
          subtext = sprintf("Baseline = %.1f (mean of 1999-to-present monthly observations)", baseline),
          left = "center") |>
  apply_econ_theme()
```

## World events observed {#events}

Events marked on the timeline above. Add more by editing
`data/sentiment/events.csv` in the repository.

```{r events-table}
#| echo: false
library(reactable)
reactable(
  events,
  striped = TRUE, compact = TRUE, highlight = TRUE,
  theme = reactableTheme(borderColor = "#dee2e6"),
  defaultSorted = list(date = "asc"),
  columns = list(
    date     = colDef(name = "Date", maxWidth = 130),
    category = colDef(name = "Category", maxWidth = 140),
    event    = colDef(name = "Event", minWidth = 380),
    severity = colDef(name = "Severity", maxWidth = 100)
  )
)
```

## Source

- **University of Michigan Surveys of Consumers**, distributed via FRED:
  *University of Michigan: Consumer Sentiment* [UMCSENT].
  <https://fred.stlouisfed.org/series/UMCSENT> — public-domain.
