---
title: "Public Presidential Sentiment"
subtitle: "Approval, economic sentiment, and media sentiment by administration — 1999 to present"
---

```{r setup}
#| include: false
source("R/helpers.R")
source("R/presidential_helpers.R")

sent <- load_admin_sentiment()
baseline_row <- sent |> filter(is.na(party))
sent <- sent |> filter(!is.na(party))

sent <- sent |>
  mutate(
    label       = admin_short_label(president, ongoing),
    party_color = party_color(party)
  )

umcsent_baseline <- baseline_row$umcsent_avg[1]
gallup_baseline  <- baseline_row$gallup_avg[1]

color_array_js <- sprintf("[%s]",
  paste(sprintf('"%s"', sent$party_color), collapse = ","))
bar_color_js <- htmlwidgets::JS(sprintf(
  "function(p){ var c=%s; return c[p.dataIndex]; }", color_array_js))
```

This report measures three different dimensions of public sentiment during
each presidential administration from 1999 to the present:

1. **Political approval** — aggregate Gallup approval averages by administration
   (hand-curated; polling APIs aren't free for full history).
2. **Economic sentiment** — University of Michigan Consumer Sentiment Index
   (`UMCSENT`, monthly, via FRED — public-domain; auto-refreshed daily).
3. **Media sentiment** — GDELT Doc 2.0 daily average-tone of US-presidency
   English-language news (CC-BY-NC; 2017-present only).

::: {.callout-important appearance="simple"}
None of these measures capture presidents directly — they capture **public
response** during a period that includes the president plus every other
influence on public mood (wars, pandemics, the business cycle, global shocks,
media environment). Treat these as descriptive, not causal. See
[Methodology](about.qmd) for the full caveats.
:::

## Baselines

Across the administrations displayed (Clinton partial → Trump II):

- **UMCSENT baseline:** `r umcsent_baseline` (mean of monthly observations
  in the 1999-to-present window that are tagged to a displayed admin).
- **Gallup baseline:** `r gallup_baseline`% (mean of the five completed
  admins' term-average approvals).

Numbers further right of zero in the "vs baseline" columns mean **above**
the cross-administration average; left of zero means **below**.

## Summary by administration {#summary}

```{r summary-table}
#| echo: false
library(reactable)
library(htmltools)

party_badge <- function(party) {
  col <- party_color(party)
  tags$span(
    style = sprintf(
      "display:inline-block; padding:2px 8px; border-radius:3px; font-size:0.72rem;
       font-weight:600; color:#fff; background:%s;", col),
    toupper(party)
  )
}

delta_style <- function(v) {
  if (is.na(v)) return(list())
  list(color = ifelse(v < 0, "#bc4749", "#2f9e44"), fontWeight = 600)
}

sent |>
  transmute(
    president = paste0(president, ifelse(ongoing, " *", "")),
    party,
    gallup_avg,
    gallup_vs_baseline,
    umcsent_avg,
    umcsent_vs_baseline
  ) |>
  reactable(
    striped = TRUE, highlight = TRUE, compact = TRUE,
    theme = reactableTheme(borderColor = "#dee2e6"),
    columns = list(
      president = colDef(name = "President", minWidth = 180),
      party = colDef(name = "Party", minWidth = 110,
                     cell = function(v) party_badge(v), html = TRUE),
      gallup_avg = colDef(name = "Gallup avg (%)",
                          format = colFormat(digits = 1, suffix = "%")),
      gallup_vs_baseline = colDef(name = "vs baseline",
                                  format = colFormat(digits = 1),
                                  style = function(v) delta_style(v)),
      umcsent_avg = colDef(name = "UMCSENT avg",
                           format = colFormat(digits = 1)),
      umcsent_vs_baseline = colDef(name = "vs baseline",
                                   format = colFormat(digits = 1),
                                   style = function(v) delta_style(v))
    )
  )
```

## Political approval vs economic sentiment {#combined}

Two different things the two measures capture: approval is *about the president*,
UMCSENT is *about the economy the public perceives*. They often move together
but don't have to.

```{r combined-bars}
#| echo: false
sent |>
  e_charts(label) |>
  e_bar(gallup_avg, name = "Gallup approval avg (%)",
        itemStyle = list(color = bar_color_js)) |>
  e_bar(umcsent_avg, name = "UMCSENT avg", y_index = 1,
        itemStyle = list(color = "#6a4c93"), barGap = "20%") |>
  e_y_axis(name = "Approval (%)", axisLabel = list(formatter = "{value}%"),
           max = 80) |>
  e_y_axis(index = 1, name = "UMCSENT", position = "right", max = 120) |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? v.toFixed(1) : v; }")) |>
  e_legend(top = 24) |>
  e_title("Gallup approval (bars) vs UMCSENT (overlay)",
          subtext = "Two separate measures of public mood; dual y-axes",
          left = "center") |>
  apply_econ_theme()
```

## Where to go next {#next}

- **[Political Approval →](approval.qmd)** — Gallup per-admin averages, min/max volatility, exit vs. peak comparison.
- **[Economic Sentiment →](economic.qmd)** — UMCSENT timeline shaded by administration, world events overlay, per-admin averages vs long-run baseline.
- **[Media Sentiment →](media.qmd)** — GDELT-derived average-tone timeline, per-admin tone averages, coverage begins 2017.
- **[Methodology →](about.qmd)** — source provenance, why "sentiment toward an administration" is imprecise, caveats.
