---
title: "Media Sentiment"
subtitle: "GDELT average-tone of US-presidency English-language news, 2017 to present"
---

```{r setup}
#| include: false
source("R/helpers.R")
source("R/presidential_helpers.R")

sent    <- load_admin_sentiment() |> filter(!is.na(party))
monthly <- read_csv(file.path(DATA_ROOT, "sentiment", "gdelt_tone_monthly.csv"),
                    show_col_types = FALSE) |>
             mutate(date = as.Date(date))
events  <- load_world_events()

baseline_row <- load_admin_sentiment() |> filter(is.na(party))
tone_baseline <- baseline_row$tone_avg[1]

# Media Cloud monthly volume (may be empty if Media Cloud cache not yet populated)
mc_path <- file.path(DATA_ROOT, "sentiment", "mediacloud_volume_monthly.csv")
mc_monthly <- if (file.exists(mc_path)) {
  read_csv(mc_path, show_col_types = FALSE) |> mutate(date = as.Date(date))
} else {
  tibble::tibble(date = as.Date(character()), mc_relevant = integer(),
                 mc_total = integer(), president = character(), party = character())
}
mc_available <- nrow(mc_monthly) > 0

# Drop admins with zero / negligible coverage from the bar chart (keep in table)
sent_tone <- sent |>
  mutate(
    label       = admin_short_label(president, ongoing),
    party_color = party_color(party)
  ) |>
  filter(!is.na(tone_avg))

has_data <- sent_tone |> filter(tone_months >= 3)

sent_vol <- sent |>
  mutate(
    label       = admin_short_label(president, ongoing),
    party_color = party_color(party)
  ) |>
  filter(!is.na(mc_stories_per_month_avg))
```

::: {.callout-note appearance="simple"}
**Data source:** [GDELT](https://www.gdeltproject.org/) Doc 2.0 API — daily
average *tone* of English-language news articles from US sources matching
`("white house" OR president OR presidential)`. Licensed **CC-BY-NC**:
redistributed here for non-commercial analytical purposes with attribution.
Not affiliated with or endorsed by The GDELT Project.

**Coverage:** For this query, GDELT returned data starting **January 2017**.
Administrations before Trump I (Clinton, Bush 43, Obama except his final
month) are absent from this section and show no value.
:::

## What "tone" is, and isn't

GDELT computes tone as the difference between positive- and negative-valence
word counts in each article, normalized to roughly **−100 to +100**. Typical
US political news sits in the **−1 to −3** range — slightly negative on
average, which is consistent with news writing conventions (crises and
conflicts drive more coverage than stability).

Tone is not the same as approval:

- It measures the **tone of articles**, not the **opinion of readers**.
- A very positive article can be written about a policy many readers
  oppose; a very negative article can cover a president's political
  opponents.
- A lexicon-based score cannot distinguish subject from context: "Trump
  condemned the attack" is neutral about Trump but GDELT's tone picks up
  the word "attack."

Treat this as a **coarse environmental signal** about the media climate
during each administration, not a precision instrument.

## Monthly tone — by administration {#timeline}

Chart is shaded by administration. Vertical markers are major world events
(same list used on the [Economic Sentiment](economic.qmd) page). Horizontal
dotted line is the long-run baseline (`r round(tone_baseline, 2)`).

```{r timeline}
#| echo: false
admins_ref <- load_administrations() |>
  mutate(start_date = as.Date(start_date),
         effective_end = as.Date(effective_end))

gdelt_first <- min(monthly$date)

bands <- admins_ref |>
  mutate(
    color = party_color(party),
    label = admin_short_label(president, ongoing),
    clip_start = pmax(start_date, gdelt_first),
    clip_end   = pmin(effective_end, max(monthly$date))
  ) |>
  filter(clip_end > clip_start)

mark_bands <- lapply(seq_len(nrow(bands)), function(i) list(
  list(xAxis = format(bands$clip_start[i], "%Y-%m-%d"),
       itemStyle = list(color = paste0(bands$color[i], "22")),
       name = bands$label[i]),
  list(xAxis = format(bands$clip_end[i],   "%Y-%m-%d"))
))

event_lines <- lapply(seq_len(nrow(events)), function(i) list(
  xAxis = format(events$date[i], "%Y-%m-%d"),
  label = list(formatter = events$event[i], position = "insideEndTop",
               rotate = 90, fontSize = 9, color = "#6c757d"),
  lineStyle = list(color = "#495057", type = "dashed", width = 1, opacity = 0.6)
))

monthly |>
  e_charts(date) |>
  e_line(tone, name = "Avg tone (GDELT)", symbol = "circle",
         symbolSize = 5, smooth = TRUE,
         lineStyle = list(width = 2, color = "#1d3557")) |>
  e_mark_area(data = mark_bands, silent = TRUE,
              label = list(show = TRUE, position = "insideTop",
                           color = "#495057", fontSize = 10)) |>
  e_mark_line(data = event_lines, symbol = "none", silent = TRUE) |>
  e_mark_line(data = list(yAxis = tone_baseline),
              lineStyle = list(type = "dotted", color = "#bc4749"),
              label = list(formatter = sprintf("Baseline %.2f", tone_baseline),
                           position = "end")) |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? v.toFixed(2) : v; }")) |>
  e_datazoom(x_index = 0) |>
  e_y_axis(name = "Avg tone") |>
  e_title("GDELT monthly average tone — US-presidency English news",
          left = "center") |>
  apply_econ_theme()
```

## Per-administration average tone {#avg}

Only administrations with at least 3 months of GDELT coverage are charted.

```{r admin-bar}
#| echo: false
color_array_js <- sprintf("[%s]",
  paste(sprintf('"%s"', has_data$party_color), collapse = ","))
bar_color_js <- htmlwidgets::JS(sprintf(
  "function(p){ var c=%s; return c[p.dataIndex]; }", color_array_js))

has_data |>
  e_charts(label) |>
  e_bar(tone_avg, name = "Avg tone",
        itemStyle = list(color = bar_color_js)) |>
  e_mark_line(data = list(yAxis = tone_baseline),
              lineStyle = list(type = "dashed", color = "#6c757d"),
              label = list(formatter = sprintf("Baseline %.2f", tone_baseline),
                           position = "end")) |>
  e_y_axis(name = "Avg tone") |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? v.toFixed(2) : v; }")) |>
  e_title("Term-average media tone per administration",
          left = "center") |>
  apply_econ_theme()
```

## Vs baseline {#vs-baseline}

```{r vs-baseline}
#| echo: false
has_data |>
  e_charts(label) |>
  e_bar(tone_vs_baseline, name = "vs baseline",
        itemStyle = list(color = bar_color_js)) |>
  e_mark_line(data = list(yAxis = 0), lineStyle = list(color = "#6c757d"),
              symbol = "none") |>
  e_y_axis(name = "Tone points") |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? (v > 0 ? '+' : '')+v.toFixed(3) : v; }")) |>
  e_title("Media tone minus long-run baseline",
          subtext = sprintf("Baseline = %.2f (mean of monthly tone, 2017-to-present)", tone_baseline),
          left = "center") |>
  apply_econ_theme()
```

## Full table

```{r full-table}
#| echo: false
library(reactable); library(htmltools)

party_badge <- function(party) {
  if (is.na(party)) return(tags$span(""))
  col <- party_color(party)
  tags$span(
    style = sprintf(
      "display:inline-block; padding:2px 8px; border-radius:3px; font-size:0.72rem;
       font-weight:600; color:#fff; background:%s;", col),
    toupper(party)
  )
}

sent |>
  mutate(president = paste0(president, ifelse(ongoing, " *", ""))) |>
  transmute(president, party, tone_avg, tone_min, tone_max, tone_months,
            tone_vs_baseline) |>
  reactable(
    striped = TRUE, highlight = TRUE, compact = TRUE,
    theme = reactableTheme(borderColor = "#dee2e6"),
    columns = list(
      president = colDef(name = "President", minWidth = 180),
      party = colDef(name = "Party", minWidth = 110,
                     cell = function(v) party_badge(v), html = TRUE),
      tone_avg = colDef(name = "Tone avg",
                        format = colFormat(digits = 2),
                        style = list(fontWeight = 600)),
      tone_min = colDef(name = "Min",  format = colFormat(digits = 2)),
      tone_max = colDef(name = "Max",  format = colFormat(digits = 2)),
      tone_months = colDef(name = "Months with data", maxWidth = 160),
      tone_vs_baseline = colDef(name = "vs baseline",
                                format = colFormat(digits = 2),
                                style = function(v) {
                                  if (is.na(v)) return(list())
                                  list(color = ifelse(v < 0, "#bc4749", "#2f9e44"),
                                       fontWeight = 600)
                                })
    )
  )
```

## Media attention — story volume (Media Cloud) {#volume}

```{r volume}
#| echo: false
#| eval: !expr mc_available
mc_monthly |>
  e_charts(date) |>
  e_line(mc_relevant, name = "Matching stories / month",
         symbol = "none", smooth = TRUE,
         lineStyle = list(width = 2, color = "#6a4c93")) |>
  e_mark_area(data = mark_bands, silent = TRUE,
              label = list(show = TRUE, position = "insideTop",
                           color = "#495057", fontSize = 10)) |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? v.toLocaleString() : v; }")) |>
  e_datazoom(x_index = 0) |>
  e_y_axis(name = "Stories") |>
  e_title("Monthly US news volume about the presidency (Media Cloud)",
          subtext = 'Matching query: ("white house" OR president OR presidential); index grows over time',
          left = "center") |>
  apply_econ_theme()
```

```{r volume-unavailable}
#| echo: false
#| eval: !expr '!mc_available'
htmltools::div(
  class = "callout callout-note",
  htmltools::p("Media Cloud volume data will appear here once the cache is populated. ",
               "The CI workflow runs the Media Cloud fetcher daily; if you just deployed, ",
               "give it a cycle to refresh.")
)
```

## Per-admin media attention — avg stories / month (Media Cloud)

```{r volume-bar}
#| echo: false
#| eval: !expr 'nrow(sent_vol) > 0'
color_array_vol <- sprintf("[%s]",
  paste(sprintf('"%s"', sent_vol$party_color), collapse = ","))
bar_color_vol <- htmlwidgets::JS(sprintf(
  "function(p){ var c=%s; return c[p.dataIndex]; }", color_array_vol))

sent_vol |>
  e_charts(label) |>
  e_bar(mc_stories_per_month_avg, name = "Avg stories / month",
        itemStyle = list(color = bar_color_vol)) |>
  e_y_axis(name = "Stories / month") |>
  e_tooltip(trigger = "axis",
            valueFormatter = htmlwidgets::JS("function(v){ return typeof v==='number' ? v.toLocaleString() : v; }")) |>
  e_title("Average monthly Media Cloud story volume per administration",
          subtext = "Raw Media Cloud counts — not corpus-normalized; later years have more indexed sources",
          left = "center") |>
  apply_econ_theme()
```

## Caveats

1. **Coverage starts 2017.** GDELT's GKG indexes articles globally, but for
   this specific query the API returned daily tone data beginning in
   January 2017. Pre-2017 admins show NA.

2. **Tone is lexicon-based.** GDELT counts positive and negative valence
   words in each article. It does not understand *who the article is about*
   or *what the author's intent is*. A headline like "Biden condemns Putin"
   registers as negative because of "condemns," even though Biden is the
   subject speaking against a negative event.

3. **US-English-only slice.** The query filters to `sourcelang:eng
   sourcecountry:US`. International and non-English coverage is excluded —
   which matters because global news framing often differs from US framing.

4. **Query is a keyword proxy.** `"white house" OR president OR
   presidential` catches most US executive-branch coverage but is not an
   exact match for "articles about the sitting president."

5. **Volume varies by era.** Later years have denser news indexing, so the
   stability of averages is not uniform. The `tone_months` column in the
   table shows effective sample size per admin.

## Attribution & license

### GDELT — tone

Tone data sourced from the **GDELT Project** (*Global Database of Events,
Language, and Tone*). The GDELT project is a collaboration of Google
Jigsaw with founders Kalev Leetaru and Philip Schrodt. Data made available
under **Creative Commons Attribution-NonCommercial (CC-BY-NC)**. See
<https://www.gdeltproject.org/>.

### Media Cloud — volume

Story volume data sourced from **Media Cloud**, an open-source media
research platform operated by a consortium including the MIT Center for
Civic Media and the Berkman Klein Center for Internet & Society. Counts
retrieved from the Media Cloud Search API (total-count endpoint). Media
Cloud data is made available for academic research under their terms of
service: <https://mediacloud.org/>.

### Non-commercial use

This site's use of all media-analytics data (GDELT, Media Cloud) is
non-commercial: the site carries no ads, sells no products, and is
published for public educational research.

## Where to go next

- **[Economic Sentiment →](economic.qmd)** — UMCSENT consumer-mood timeline.
- **[Political Approval →](approval.qmd)** — Gallup per-admin averages.
- **[Methodology →](about.qmd)** — full caveats across all three sections.
