---
title: "Society Radar"
subtitle: "Sentiment across six aspects of public life, by administration"
---

```{r setup}
#| include: false
source("R/helpers.R")
source("R/presidential_helpers.R")

scores <- read_csv(file.path(DATA_ROOT, "sentiment", "society_scores.csv"),
                   show_col_types = FALSE)

# Locked aspect order around the radar (clockwise from 12 o'clock)
aspect_order <- c("Public Safety", "Economy", "Health",
                  "Prosperity", "Happiness", "Peace")

# Admin order and distinct colors (within-party shades so overlapping
# lines remain visually distinguishable)
admin_order <- c("Bill Clinton", "George W. Bush", "Barack Obama",
                 "Donald Trump (1st term)", "Joe Biden",
                 "Donald Trump (2nd term)")

admin_colors <- c(
  "Bill Clinton"             = "#003049",
  "George W. Bush"           = "#780000",
  "Barack Obama"             = "#2a6f97",
  "Donald Trump (1st term)"  = "#bc4749",
  "Joe Biden"                = "#669bbc",
  "Donald Trump (2nd term)"  = "#d62828"
)

hex_alpha <- function(hex, alpha_hex = "33") paste0(hex, alpha_hex)  # ~20% fill
```

Each line is one presidential administration. Angles are six aspects of
public life; the radial distance represents sentiment from **−10** (extreme
negative) at the center to **+10** (extreme positive) at the outer edge,
with **0** as the baseline at the middle ring. A line passing *through* the
middle ring on an axis means sentiment on that aspect was neutral during
that administration.

::: {.callout-important appearance="simple"}
**These scores are editorial calibrations.** They anchor on measurable
signals used elsewhere on this site (UMCSENT, Gallup approval, NBER
recessions, wars-in-progress, published violent-crime indices) plus
historical commentary, but the final single-number −10/+10 mapping per
aspect is a single author's aggregation. See [Methodology](about.qmd) for
the full caveats, and `data/sentiment/society_scores.csv` in the repo for
the rationale attached to each score. Corrections welcome via the repo.
:::

## Radar {#radar}

```{r radar}
#| echo: false
scores_ordered <- scores |>
  mutate(aspect = factor(aspect, levels = aspect_order),
         president = factor(president, levels = admin_order)) |>
  arrange(aspect, president)

scores_wide <- scores_ordered |>
  select(president, aspect, score) |>
  pivot_wider(names_from = president, values_from = score) |>
  arrange(aspect)

# Aspect axis indicators (same min/max for all six)
indicators <- lapply(aspect_order, function(a) {
  list(name = a, max = 10, min = -10)
})

chart <- scores_wide |> e_charts(aspect)

for (admin_name in admin_order) {
  if (!(admin_name %in% colnames(scores_wide))) next
  col <- admin_colors[[admin_name]]
  chart <- chart |>
    e_radar_(admin_name,
             max = 10, min = -10,
             name = admin_name,
             symbol = "circle", symbolSize = 6,
             itemStyle = list(color = col),
             lineStyle = list(color = col, width = 2),
             areaStyle = list(color = hex_alpha(col)))
}

chart <- chart |>
  e_radar_opts(
    indicator   = indicators,
    radius      = "68%",
    shape       = "polygon",
    splitNumber = 4,
    name        = list(textStyle = list(fontSize = 13, color = "#1d3557",
                                        fontWeight = 600)),
    axisLine    = list(lineStyle = list(color = "#adb5bd")),
    splitLine   = list(lineStyle = list(color = "#dee2e6")),
    splitArea   = list(show = TRUE,
                       areaStyle = list(color = c("#ffffff", "#f8f9fa")))
  ) |>
  e_tooltip(trigger = "item",
            formatter = htmlwidgets::JS(sprintf("
              function(p){
                var aspects = %s;
                var v = p.value;
                return '<b>' + p.name + '</b><br/>' +
                  aspects.map(function(a,i){
                    var val = v[i];
                    if (typeof val !== 'number' || !isFinite(val)) return a + ': —';
                    var sign = val > 0 ? '+' : '';
                    return a + ': ' + sign + val;
                  }).join('<br/>');
              }",
              jsonlite::toJSON(aspect_order)))) |>
  e_legend(top = 0, textStyle = list(fontSize = 11)) |>
  e_title("Sentiment across six aspects of society, by administration",
          subtext = "−10 extreme negative · 0 baseline (middle ring) · +10 extreme positive",
          left = "center") |>
  apply_econ_theme()

chart
```

## Scores table {#table}

Full list of curated scores, with per-row justification notes. Edit
`data/sentiment/society_scores.csv` to refine.

```{r scores-table}
#| echo: false
library(reactable); library(htmltools)

party_badge <- function(party) {
  col <- party_color(party)
  tags$span(
    style = sprintf(
      "display:inline-block; padding:2px 8px; border-radius:3px;
       font-size:0.7rem; font-weight:600; color:#fff; background:%s;", col),
    toupper(party)
  )
}

score_cell <- function(v) {
  if (is.na(v)) return("—")
  col <- if (v > 0) "#2f9e44" else if (v < 0) "#bc4749" else "#6c757d"
  sign <- if (v > 0) "+" else ""
  htmltools::div(
    style = sprintf("font-weight:600; color:%s;", col),
    paste0(sign, v)
  )
}

scores |>
  transmute(
    president = paste0(president, ifelse(
      grepl("2nd term", president), " *", "")),
    party, aspect, score, notes
  ) |>
  reactable(
    striped = TRUE, highlight = TRUE, compact = TRUE,
    theme = reactableTheme(borderColor = "#dee2e6"),
    groupBy = "president",
    defaultPageSize = 40,
    columns = list(
      president = colDef(name = "President", minWidth = 180),
      party = colDef(name = "Party", minWidth = 110,
                     cell = function(v) party_badge(v), html = TRUE,
                     aggregate = "unique"),
      aspect = colDef(name = "Aspect", minWidth = 130),
      score = colDef(name = "Score (-10…+10)", maxWidth = 130,
                     cell = score_cell, html = TRUE, align = "center",
                     aggregate = "mean",
                     format = list(aggregated = colFormat(digits = 1))),
      notes = colDef(name = "Calibration notes", minWidth = 320)
    )
  )
```

## Reading the chart

A line that stays near the **middle ring** across all six angles = a
baseline administration (nothing notably better or worse than average). A
line that spikes **outward** on one axis = notably positive on that aspect;
a line that collapses **toward the center** = notably negative.

Because the six aspects are independent, an administration can be strong on
some and weak on others — Clinton's radar fills the outer half on economy
and prosperity but sits near the middle on peace; Bush 43's inner lobe on
peace reflects the wars in progress throughout his term; Trump I's health
score collapses to the inner edge because of COVID-19's first-wave
fatalities.

## Methodology summary

- **Six aspects** chosen for their independent signal: Public Safety,
  Economy, Health, Prosperity, Happiness, Peace.
- **Scale** is −10 to +10, calibrated so that 0 = long-run baseline for US
  conditions since ~1990. Values outside ±5 are reserved for historically
  extreme situations (war, pandemic, recession depth).
- **Data anchors** per aspect:
  - *Public Safety*: violent-crime rate trend (FBI UCR); high-salience
    events; Gallup crime-worry polling.
  - *Economy*: UMCSENT, NBER recession dating, GDP growth pattern.
  - *Health*: life-expectancy trajectory, major public-health events,
    major legislative changes.
  - *Prosperity*: real-wage growth, household-wealth trajectory, inflation
    impact on cost of living.
  - *Happiness*: UMCSENT level, Gallup life-satisfaction, approval bottoms.
  - *Peace*: number and scale of active wars involving US forces; major
    international crises on US agenda.
- **Ongoing administration (Trump 2nd term)** gets partial scores; aspects
  that can't be judged yet are omitted from radar display but remain in the
  table.

See the full [Methodology page](about.qmd) for the cross-section caveats
that apply across all of Public Presidential Sentiment.

## Sources

Anchor data comes from the same sources already cited elsewhere — see
[Data & Citations](../economy/about.qmd) and [Methodology](about.qmd).
