---
title: "Botnet Hosts"
subtitle: "Active C2 infrastructure from Abuse.ch FeodoTracker, past 24 hrs"
---

```{r setup}
#| include: false
source("R/helpers.R")
DATA_CYBER <- file.path(DATA_ROOT, "cybersecurity")

read_csv_safe <- function(p) {
  if (!file.exists(p) || file.size(p) == 0) return(tibble::tibble())
  suppressWarnings(readr::read_csv(p, show_col_types = FALSE))
}

botnets       <- read_csv_safe(file.path(DATA_CYBER, "current_botnets.csv"))
malware_daily <- read_csv_safe(file.path(DATA_CYBER, "malware_family_daily.csv"))

have_data <- nrow(botnets) > 0
have_geo  <- have_data && any(!is.na(botnets$lat))
```

::: {.callout-note appearance="simple"}
**Attribution.** Botnet C2 data from
[Abuse.ch FeodoTracker](https://feodotracker.abuse.ch/) (CC0). FeodoTracker
specifically tracks command-and-control servers used by major banking-trojan
and ransomware-loader botnets (Emotet, Dridex, TrickBot, Qakbot, IcedID,
BazarLoader, etc.). IP→province geolocation by
[ip-api.com](https://ip-api.com/).
:::

## World map of active botnet C2s {#map}

```{r map}
#| echo: false
#| eval: !expr have_geo
suppressPackageStartupMessages(library(echarts4r))

mapdata <- botnets |>
  dplyr::filter(!is.na(lat), !is.na(lon)) |>
  dplyr::mutate(
    label = paste0(ifelse(is.na(city), "", paste0(city, ", ")),
                   ifelse(is.na(region_name), "", paste0(region_name, ", ")),
                   country)
  )

mapdata |>
  e_charts(lon) |>
  e_geo(
    roam = TRUE,
    itemStyle = list(areaColor = "#f8f9fa", borderColor = "#adb5bd"),
    emphasis  = list(itemStyle = list(areaColor = "#dee2e6"))
  ) |>
  e_effect_scatter(lat, coord_system = "geo", symbol_size = 8,
                   itemStyle = list(color = "#bc4749")) |>
  e_tooltip(formatter = htmlwidgets::JS(sprintf("
    function(p){
      var labels  = %s;
      var malware = %s;
      var status  = %s;
      var i = p.dataIndex;
      return labels[i] + '<br/>Malware: ' + (malware[i] || '—') +
             '<br/>Status: ' + (status[i] || '—');
    }",
    jsonlite::toJSON(mapdata$label),
    jsonlite::toJSON(mapdata$malware_family),
    jsonlite::toJSON(mapdata$status)
  ))) |>
  e_title(sprintf("FeodoTracker — %d active C2 servers across %d provinces",
                  nrow(mapdata),
                  dplyr::n_distinct(paste0(mapdata$country, "|", mapdata$region_name)))) |>
  apply_econ_theme()
```

```{r no-data}
#| echo: false
#| eval: !expr '!have_geo'
htmltools::div(class = "callout callout-note",
  htmltools::p("Botnet data isn't populated yet. CI pipeline will fetch on next run."))
```

## Malware family breakdown {#families}

```{r family-bar}
#| echo: false
#| eval: !expr have_data
suppressPackageStartupMessages(library(echarts4r))

fam <- botnets |>
  dplyr::filter(!is.na(malware_family), nzchar(malware_family)) |>
  dplyr::count(malware_family, sort = TRUE) |>
  utils::head(15)

fam |>
  e_charts(malware_family) |>
  e_bar(n, name = "Active C2 IPs",
        itemStyle = list(color = "#6a4c93")) |>
  e_x_axis(axisLabel = list(rotate = 30, interval = 0)) |>
  e_y_axis(name = "C2 IPs") |>
  e_tooltip(trigger = "axis") |>
  e_title("Active botnet C2 IPs by malware family", left = "center") |>
  e_legend(show = FALSE) |>
  apply_econ_theme()
```

## Online vs offline {#status}

FeodoTracker labels each tracked C2 as `online` (currently reachable) or
`offline` (recently seen but not currently responding).

```{r status-table}
#| echo: false
#| eval: !expr have_data
status <- botnets |>
  dplyr::filter(!is.na(status), nzchar(status)) |>
  dplyr::count(status, sort = TRUE)

suppressPackageStartupMessages(library(reactable))
reactable(status,
  striped = TRUE, highlight = TRUE, compact = TRUE,
  theme = reactableTheme(borderColor = "#dee2e6"),
  columns = list(
    status = colDef(name = "Status", minWidth = 140),
    n      = colDef(name = "Count", maxWidth = 120,
                    style = list(fontWeight = 600))
  )
)
```

## Top hosting networks for botnet C2s {#asn}

```{r asn-bot}
#| echo: false
#| eval: !expr have_data
asn <- botnets |>
  dplyr::filter(!is.na(as_name), nzchar(as_name)) |>
  dplyr::count(as_name, sort = TRUE) |>
  utils::head(15)

reactable(asn,
  striped = TRUE, highlight = TRUE, compact = TRUE,
  theme = reactableTheme(borderColor = "#dee2e6"),
  defaultPageSize = 15,
  columns = list(
    as_name = colDef(name = "Autonomous System", minWidth = 320),
    n       = colDef(name = "Active botnet C2s", maxWidth = 160,
                     style = list(fontWeight = 600))
  )
)
```
