---
title: "CVEs in the Wild"
subtitle: "CISA KEV catalog with initial vs current EPSS, CVSS, and exploit confirmation"
---

```{r setup}
#| include: false
source("R/helpers.R")
DATA_CYBER <- file.path(DATA_ROOT, "cybersecurity")

read_csv_safe <- function(p) {
  if (!file.exists(p) || file.size(p) == 0) return(tibble::tibble())
  suppressWarnings(readr::read_csv(p, show_col_types = FALSE))
}

cves         <- read_csv_safe(file.path(DATA_CYBER, "cves_kev.csv"))
cves_summary <- read_csv_safe(file.path(DATA_CYBER, "cves_summary.csv"))

have_data <- nrow(cves) > 0
```

::: {.callout-note appearance="simple"}
**Sources.** CVEs and "in-wild" status from
[CISA KEV Catalog](https://www.cisa.gov/known-exploited-vulnerabilities-catalog)
(US government work, public domain). EPSS (Exploit Prediction Scoring
System) from [FIRST EPSS](https://www.first.org/epss/) (CC-BY-SA). CVSS
scores from the [NVD](https://nvd.nist.gov/) (US government work, public
domain).

**EPSS** estimates the probability a CVE will be exploited in the next 30
days. **CVSS** is a static severity score for the vulnerability itself.
**KEV** confirms an exploit has been observed in the wild — every row on
this page is "in wild = TRUE."
:::

```{r summary-cards}
#| echo: false
#| eval: !expr 'nrow(cves_summary) > 0'
card_row(
  value_card("KEV CVEs total",
             format(cves_summary$kev_total, big.mark = ","),
             sprintf("As of %s", cves_summary$as_of),
             tone = "negative"),
  value_card("New in last 30 days",
             format(cves_summary$kev_added_30d, big.mark = ","),
             "Newly added by CISA",
             tone = "warn"),
  value_card("Top CVE by current EPSS",
             cves_summary$top_epss,
             sprintf("EPSS = %.3f", cves_summary$top_epss_score),
             tone = "negative"),
  value_card("Median current EPSS (KEV)",
             sprintf("%.3f", cves_summary$median_epss),
             "Across all KEV-listed CVEs",
             tone = "warn"),
  value_card("Median CVSS v3 (KEV)",
             sprintf("%.1f", cves_summary$median_cvss_v3),
             "Severity of the vuln itself",
             tone = "warn")
)
```

## All KEV CVEs — sortable, searchable {#table}

EPSS values range 0.0 – 1.0 (probability of exploitation in next 30 days).
CVSS v3 ranges 0.0 – 10.0 (severity). The **EPSS Δ** column shows how a
CVE's predicted exploit probability has moved since we first observed it
in our cache.

```{r cve-table}
#| echo: false
#| eval: !expr have_data
suppressPackageStartupMessages({ library(reactable); library(htmltools) })

severity_pill <- function(sev) {
  if (is.na(sev)) return(htmltools::span("—"))
  col <- switch(toupper(sev),
                CRITICAL = "#7f1d1d", HIGH = "#bc4749",
                MEDIUM = "#d68102", LOW = "#2f9e44",
                "#6c757d")
  htmltools::span(
    style = sprintf(
      "display:inline-block; padding:1px 7px; border-radius:3px;
       font-size:0.7rem; font-weight:600; color:#fff; background:%s;",
      col),
    toupper(sev)
  )
}

epss_bar <- function(v) {
  if (is.na(v)) return("—")
  pct <- round(v * 100)
  width <- paste0(pct, "%")
  col <- if (pct >= 90) "#7f1d1d"
         else if (pct >= 50) "#bc4749"
         else if (pct >= 10) "#d68102"
         else "#2f9e44"
  htmltools::tagList(
    htmltools::div(
      style = "display:flex; align-items:center; gap:6px;",
      htmltools::div(
        style = "flex:1; background:#e9ecef; height:6px; border-radius:3px; min-width:50px;",
        htmltools::div(style = sprintf("background:%s; height:6px; width:%s; border-radius:3px;",
                                       col, width))),
      htmltools::div(style = "width:42px; text-align:right; font-variant-numeric:tabular-nums;",
                     sprintf("%.3f", v))
    )
  )
}

delta_cell <- function(v) {
  if (is.na(v)) return("—")
  if (abs(v) < 0.001) return(htmltools::span(style = "color:#6c757d", "0"))
  col <- if (v > 0) "#bc4749" else "#2f9e44"
  sign <- if (v > 0) "▲" else "▼"
  htmltools::span(style = sprintf("color:%s; font-weight:600;", col),
                  sprintf("%s %.3f", sign, abs(v)))
}

cve_link <- function(id) htmltools::a(href = paste0("https://nvd.nist.gov/vuln/detail/", id),
                                       target = "_blank", id)

cves |>
  dplyr::transmute(
    cve, vendor_project, product, vulnerability_name,
    cvss_v3_base, cvss_v3_severity,
    initial_epss, current_epss, epss_delta,
    date_added
  ) |>
  reactable(
    striped = TRUE, highlight = TRUE, compact = TRUE, searchable = TRUE,
    filterable = FALSE, defaultPageSize = 25, showPageSizeOptions = TRUE,
    pageSizeOptions = c(10, 25, 50, 100),
    theme = reactableTheme(borderColor = "#dee2e6"),
    defaultSorted = list(current_epss = "desc"),
    columns = list(
      cve = colDef(name = "CVE", minWidth = 145,
                   cell = cve_link, html = TRUE),
      vendor_project = colDef(name = "Vendor", minWidth = 130),
      product = colDef(name = "Product", minWidth = 160),
      vulnerability_name = colDef(name = "Vulnerability", minWidth = 240),
      cvss_v3_base = colDef(name = "CVSS v3", maxWidth = 95,
                            format = colFormat(digits = 1)),
      cvss_v3_severity = colDef(name = "Severity", maxWidth = 110,
                                cell = severity_pill, html = TRUE,
                                align = "center"),
      initial_epss = colDef(name = "EPSS (initial)", minWidth = 140,
                            cell = epss_bar, html = TRUE),
      current_epss = colDef(name = "EPSS (current)", minWidth = 140,
                            cell = epss_bar, html = TRUE),
      epss_delta = colDef(name = "EPSS Δ", maxWidth = 110,
                          cell = delta_cell, html = TRUE, align = "center"),
      date_added = colDef(name = "Added to KEV", maxWidth = 130)
    )
  )
```

```{r no-data}
#| echo: false
#| eval: !expr '!have_data'
htmltools::div(class = "callout callout-note",
  htmltools::p("CVE data not yet populated. CI pipeline will populate on next ",
               "deploy. Note: NVD lookups are rate-limited; first bootstrap may ",
               "take ~10 minutes for the full KEV catalog."))
```

## How EPSS and CVSS differ

- **CVSS v3** scores severity on a 0–10 scale based on the vulnerability's
  intrinsic characteristics: attack vector, complexity, required privileges,
  user interaction, scope, and impact on confidentiality/integrity/
  availability. CVSS does not change much over time.
- **EPSS** estimates the *probability* a vulnerability will be exploited in
  the next 30 days, based on observed exploit behavior, public discussion,
  and threat-actor activity. EPSS updates daily and can move a lot — a CVE
  might have low CVSS but high EPSS (easy to exploit, attractive target),
  or vice versa.
- **KEV inclusion** is the strongest signal: it confirms exploitation has
  *already been observed* by US federal agencies. Every CVE on this page is
  in KEV.

EPSS Δ tracks **how predicted-exploit probability has moved** since this
site first observed the CVE. A large positive Δ means EPSS thinks the
likelihood went up; large negative means it went down.

## Source notes

- **CISA KEV** updates irregularly (typically several times per month).
  Refresh: daily.
- **EPSS** publishes a daily snapshot of all ~250k tracked CVEs. We retain
  the *first* score we see per CVE as `initial_epss` (immutable) and update
  `current_epss` each day.
- **NVD CVSS** is fetched on-demand for KEV CVEs we haven't yet scored,
  rate-limited to ≤2 req/sec without an API key.
