---
title: "Cybersecurity Threats"
subtitle: "Active threat infrastructure + known-exploited vulnerabilities"
---

```{r setup}
#| include: false
source("R/helpers.R")

DATA_CYBER <- file.path(DATA_ROOT, "cybersecurity")

read_csv_safe <- function(p) {
  if (!file.exists(p) || file.size(p) == 0) return(tibble::tibble())
  suppressWarnings(readr::read_csv(p, show_col_types = FALSE))
}

threats_summary <- read_csv_safe(file.path(DATA_CYBER, "threats_summary.csv"))
cves_summary    <- read_csv_safe(file.path(DATA_CYBER, "cves_summary.csv"))

`%||%` <- function(a, b) if (is.null(a) || length(a) == 0 || (length(a) == 1 && is.na(a))) b else a
```

This report tracks two complementary cybersecurity signals:

1. **Active threat infrastructure** — IP addresses currently used for malicious
   purposes (botnet C2s, malware hosts, phishing kits) drawn from public
   feeds and geolocated to the **city/province level**.
2. **Known-exploited vulnerabilities** — the CISA KEV catalog, enriched with
   FIRST EPSS scores (initial vs current) and NVD CVSS scores so you can see
   not just which CVEs are being exploited but how their predicted-exploit
   probability is moving.

Daily snapshots accumulate over time so we can show trends, not just the
"as of right now" view.

::: {.callout-note appearance="simple"}
**Hosting location ≠ attacker location.** Attackers routinely use rented
hosting in countries with weak attribution. This report shows where
malicious *infrastructure* is, not where its operators sit.
:::

## Today's snapshot {#snapshot}

```{r snapshot-cards}
#| echo: false
if (nrow(threats_summary) > 0) {
  cards <- list(
    value_card("Active threat IPs",
               format(threats_summary$total_ips_today, big.mark = ","),
               sprintf("As of %s", threats_summary$as_of),
               tone = "warn"),
    value_card("Botnet C2 hosts",
               format(threats_summary$total_botnet_ips_today, big.mark = ","),
               "FeodoTracker, currently online",
               tone = "negative"),
    value_card("Provinces affected",
               format(threats_summary$provinces_today, big.mark = ","),
               sprintf("Across %d countries", threats_summary$countries_today),
               tone = "neutral"),
    value_card("Top malware family",
               threats_summary$top_malware_today %||% "—",
               "Most-reported in today's snapshot",
               tone = "warn"),
    value_card("Daily snapshots stored",
               format(threats_summary$snapshots_accumulated, big.mark = ","),
               "Each one a 24-hour view; cache grows daily",
               tone = "neutral")
  )
  if (nrow(cves_summary) > 0) {
    cards <- c(cards, list(
      value_card("KEV CVEs (in-wild)",
                 format(cves_summary$kev_total, big.mark = ","),
                 sprintf("+%d added in last 30 days", cves_summary$kev_added_30d),
                 tone = "negative")
    ))
  }
  do.call(card_row, cards)
} else {
  htmltools::div(class = "callout callout-warning",
    htmltools::p("Cybersecurity data is not yet populated. The CI pipeline ",
                 "fetches it on first run; check back after the next deploy."))
}
```

## What's in here

- **[Threat sources →](threats.qmd)** — World map of every active threat IP
  geolocated to its city/province; province aggregation table; top ASNs.
- **[Botnet hosts →](botnets.qmd)** — FeodoTracker C2 infrastructure
  specifically; malware-family breakdown; online vs offline.
- **[CVEs in the wild →](cves.qmd)** — CISA KEV catalog with initial vs
  current EPSS, CVSS v3, and "in wild" flag. Sorted by current EPSS.
- **[Methodology →](about.qmd)** — Sources, attribution, refresh cadence,
  caveats.

## Update cadence

The CI pipeline fetches all four feeds **once daily**:

| Feed | What | License | Cadence |
|---|---|---|---|
| Abuse.ch FeodoTracker | Active botnet C2 IPs | CC0 | Continuous; we snapshot daily |
| Abuse.ch ThreatFox | Recent IoCs incl. C2 IPs | CC0 | Recent-window CSV; snapshot daily |
| CISA KEV | Known-exploited CVEs | Public domain | When CISA updates (≈weekly) |
| FIRST EPSS | Daily exploit-probability per CVE | CC-BY-SA | Daily |
| ip-api.com | IP→province geolocation | Free non-commercial | On-demand for new IPs |

Per-IP geolocations are looked up **once** and cached forever, so daily
runs only resolve newly-seen IPs.
