---
title: "Energy Events & Sentiment"
subtitle: "Historical and recent energy-market shocks with sentiment scores"
---

```{r setup}
#| include: false
source("R/helpers.R")
ENERGY_DIR <- file.path(DATA_ROOT, "energy")

read_csv_safe <- function(p) {
  if (!file.exists(p) || file.size(p) == 0) return(tibble::tibble())
  suppressWarnings(readr::read_csv(p, show_col_types = FALSE))
}

events <- read_csv_safe(file.path(ENERGY_DIR, "events_energy.csv"))
prices <- read_csv_safe(file.path(ENERGY_DIR, "us_prices_daily.csv"))

have_events <- nrow(events) > 0
have_prices <- nrow(prices) > 0
```

::: {.callout-note appearance="simple"}
**Sentiment values are editorial.** Each score in
`data/energy/events_energy.csv` is a −1..+1 reading of contemporaneous
market reaction plus near-term policy/industry impact. These are a
single-author aggregation — the CSV's `notes` column explains the anchor
for each row. Future enhancement: derive sentiment from GDELT tone around
the event date.
:::

## Timeline: crude prices with event overlays {#timeline}

```{r timeline}
#| echo: false
#| eval: !expr 'have_events && have_prices'
suppressPackageStartupMessages(library(echarts4r))

sentimentColor <- function(s) {
  if (is.na(s)) return("#adb5bd")
  if (s >= 0.3)  return("#2f9e44")
  if (s >= 0.0)  return("#8cbf4f")
  if (s >= -0.3) return("#f2c14e")
  if (s >= -0.6) return("#e07a5f")
  "#bc4749"
}

event_lines <- lapply(seq_len(nrow(events)), function(i) list(
  xAxis = format(events$date[i], "%Y-%m-%d"),
  label = list(
    formatter = paste0(events$event[i], " (", sprintf("%+.2f", events$sentiment[i]), ")"),
    position = "insideEndTop", rotate = 90, fontSize = 9,
    color = sentimentColor(events$sentiment[i]),
    fontWeight = 600
  ),
  lineStyle = list(color = sentimentColor(events$sentiment[i]),
                   type = "dashed", width = 1, opacity = 0.75)
))

prices |>
  dplyr::mutate(date = as.Date(date)) |>
  dplyr::arrange(date) |>
  e_charts(date) |>
  e_line(wti, name = "WTI ($/bbl)", symbol = "none", smooth = TRUE,
         lineStyle = list(width = 2, color = "#1d3557")) |>
  e_mark_line(data = event_lines, symbol = "none", silent = TRUE) |>
  e_tooltip(trigger = "axis") |>
  e_datazoom(x_index = 0) |>
  e_y_axis(name = "USD / barrel") |>
  e_title("WTI crude with energy-event markers (colored by sentiment)",
          subtext = "Red = negative market impact, green = positive, amber = mixed",
          left = "center") |>
  apply_econ_theme()
```

## Recent events (past 24 months) {#recent}

```{r recent}
#| echo: false
#| eval: !expr have_events
suppressPackageStartupMessages({ library(reactable); library(htmltools) })

cutoff <- Sys.Date() - 730
recent <- events |>
  dplyr::filter(date >= cutoff) |>
  dplyr::arrange(dplyr::desc(date))

sentiment_cell <- function(v) {
  if (is.na(v)) return("—")
  col <- if (v > 0.2) "#2f9e44"
         else if (v >= 0) "#6c757d"
         else if (v >= -0.3) "#d68102"
         else "#bc4749"
  sign <- if (v > 0) "+" else ""
  htmltools::span(style = sprintf("color:%s; font-weight:600;", col),
                  sprintf("%s%.2f", sign, v))
}

if (nrow(recent) > 0) {
  reactable(recent |> transmute(date, category, event, sentiment, notes),
    striped = TRUE, highlight = TRUE, compact = TRUE,
    theme = reactableTheme(borderColor = "#dee2e6"),
    columns = list(
      date      = colDef(name = "Date", maxWidth = 120),
      category  = colDef(name = "Category", maxWidth = 140),
      event     = colDef(name = "Event", minWidth = 280),
      sentiment = colDef(name = "Sentiment", maxWidth = 110,
                         cell = sentiment_cell, html = TRUE, align = "center"),
      notes     = colDef(name = "Notes", minWidth = 320)
    )
  )
} else {
  htmltools::p("No events in the past 24 months.")
}
```

## Full historical events {#historical}

```{r historical}
#| echo: false
#| eval: !expr have_events
reactable(events |> transmute(date, category, event, sentiment, notes),
  striped = TRUE, highlight = TRUE, compact = TRUE,
  theme = reactableTheme(borderColor = "#dee2e6"),
  defaultSorted = list(date = "desc"),
  defaultPageSize = 15,
  showPageSizeOptions = TRUE,
  pageSizeOptions = c(10, 15, 25, 50),
  columns = list(
    date      = colDef(name = "Date", maxWidth = 120),
    category  = colDef(name = "Category", maxWidth = 140),
    event     = colDef(name = "Event", minWidth = 300),
    sentiment = colDef(name = "Sentiment", maxWidth = 110,
                       cell = sentiment_cell, html = TRUE, align = "center"),
    notes     = colDef(name = "Notes", minWidth = 340)
  )
)
```
