---
title: "Energy Forecasts"
subtitle: "EIA Short-Term Energy Outlook (STEO) — monthly forecast through next ~24 months"
---

```{r setup}
#| include: false
source("R/helpers.R")
ENERGY_DIR <- file.path(DATA_ROOT, "energy")

read_csv_safe <- function(p) {
  if (!file.exists(p) || file.size(p) == 0) return(tibble::tibble())
  suppressWarnings(readr::read_csv(p, show_col_types = FALSE))
}

steo   <- read_csv_safe(file.path(ENERGY_DIR, "steo_forecast.csv"))
prices <- read_csv_safe(file.path(ENERGY_DIR, "us_prices_daily.csv"))

have_steo <- nrow(steo) > 0
```

::: {.callout-note appearance="simple"}
**Source:** EIA Short-Term Energy Outlook (STEO) — published monthly,
forecast horizon ~24 months. Public domain. The long-term horizon (20+
years) is covered by EIA's Annual Energy Outlook (AEO), which we don't
publish here because AEO is annual and scenario-dependent; interested
readers should go direct to
[eia.gov/outlooks/aeo/](https://www.eia.gov/outlooks/aeo/).
:::

## WTI crude spot — history + STEO forecast {#wti-forecast}

```{r wti-forecast}
#| echo: false
#| eval: !expr 'have_steo && nrow(prices) > 0'
suppressPackageStartupMessages(library(echarts4r))

wti_hist <- prices |>
  dplyr::filter(!is.na(wti)) |>
  dplyr::transmute(date = as.Date(date), value = wti, series = "History")

wti_fore <- steo |>
  dplyr::filter(series_id == "WTIPUUS") |>
  dplyr::transmute(date = as.Date(date), value, series = "STEO forecast")

combined <- dplyr::bind_rows(wti_hist, wti_fore) |>
  dplyr::arrange(date) |>
  tidyr::pivot_wider(names_from = series, values_from = value)

combined |>
  e_charts(date) |>
  e_line(`History`, name = "WTI history", symbol = "none", smooth = TRUE,
         connectNulls = FALSE,
         lineStyle = list(width = 2, color = "#1d3557")) |>
  e_line(`STEO forecast`, name = "STEO forecast", symbol = "none", smooth = TRUE,
         connectNulls = FALSE,
         lineStyle = list(width = 2, color = "#e07a5f", type = "dashed")) |>
  e_tooltip(trigger = "axis") |>
  e_datazoom(x_index = 0, start = 70, end = 100) |>
  e_y_axis(name = "USD / bbl") |>
  e_title("WTI crude — history and STEO forecast",
          subtext = "History: FRED DCOILWTICO daily · Forecast: EIA STEO monthly",
          left = "center") |>
  apply_econ_theme()
```

## US crude production forecast {#prod-forecast}

```{r prod-forecast}
#| echo: false
#| eval: !expr have_steo
steo_prod <- steo |> dplyr::filter(series_id == "COPRPUS")

if (nrow(steo_prod) > 0) {
  steo_prod |>
    dplyr::mutate(date = as.Date(date)) |>
    dplyr::arrange(date) |>
    e_charts(date) |>
    e_line(value, name = "US crude production forecast", symbol = "none",
           smooth = TRUE, lineStyle = list(width = 2, color = "#2a6f97"),
           areaStyle = list(color = "rgba(42,111,151,0.15)")) |>
    e_tooltip(trigger = "axis") |>
    e_y_axis(name = "Million barrels / day") |>
    e_title("STEO US crude-oil production forecast", left = "center") |>
    apply_econ_theme()
} else {
  htmltools::p("No STEO production series cached.")
}
```

## Henry Hub natural gas forecast {#natgas-forecast}

```{r natgas-forecast}
#| echo: false
#| eval: !expr have_steo
steo_ng <- steo |> dplyr::filter(series_id == "NGHHMCF")

if (nrow(steo_ng) > 0) {
  steo_ng |>
    dplyr::mutate(date = as.Date(date)) |>
    dplyr::arrange(date) |>
    e_charts(date) |>
    e_line(value, name = "Henry Hub forecast", symbol = "none", smooth = TRUE,
           lineStyle = list(width = 2, color = "#6a4c93"),
           areaStyle = list(color = "rgba(106,76,147,0.12)")) |>
    e_tooltip(trigger = "axis") |>
    e_y_axis(name = "USD / MMBtu") |>
    e_title("STEO Henry Hub natural-gas forecast", left = "center") |>
    apply_econ_theme()
} else {
  htmltools::p("No STEO natural-gas series cached.")
}
```

```{r no-data}
#| echo: false
#| eval: !expr '!have_steo'
htmltools::div(class = "callout callout-warning",
  htmltools::p("STEO forecast data not yet populated — ensure EIA_API_KEY ",
               "is set and rerun the pipeline. STEO bootstrap fetches ~70 ",
               "monthly observations per series."))
```
