---
title: "International Energy Markets"
subtitle: "WTI vs Brent spread + international context"
---

```{r setup}
#| include: false
source("R/helpers.R")
ENERGY_DIR <- file.path(DATA_ROOT, "energy")

read_csv_safe <- function(p) {
  if (!file.exists(p) || file.size(p) == 0) return(tibble::tibble())
  suppressWarnings(readr::read_csv(p, show_col_types = FALSE))
}

prices <- read_csv_safe(file.path(ENERGY_DIR, "us_prices_daily.csv"))
have_prices <- nrow(prices) > 0
```

::: {.callout-note appearance="simple"}
**Retail fuel prices by country** aren't available through any free public
API — EIA doesn't publish them; IEA keeps them behind a paid subscription;
commercial feeds (Platts, Argus, IHS Markit) are licensed. This page shows
the available free-data alternative: the **WTI–Brent spread** as a proxy
for the US-vs-international crude-benchmark gap.
:::

## WTI vs Brent {#wti-brent}

WTI (US benchmark, Cushing Oklahoma) and Brent (North Sea, international
waterborne benchmark) diverge when US shale production outruns takeaway
capacity (WTI discount widens) or when geopolitical risk concentrates in
the Middle East / Europe (Brent premium widens).

```{r wti-brent}
#| echo: false
#| eval: !expr have_prices
suppressPackageStartupMessages(library(echarts4r))

prices |>
  dplyr::mutate(date = as.Date(date)) |>
  dplyr::filter(!is.na(wti), !is.na(brent)) |>
  dplyr::arrange(date) |>
  e_charts(date) |>
  e_line(wti, name = "WTI ($/bbl)", symbol = "none", smooth = TRUE,
         lineStyle = list(width = 2, color = "#2a6f97")) |>
  e_line(brent, name = "Brent ($/bbl)", symbol = "none", smooth = TRUE,
         lineStyle = list(width = 2, color = "#e07a5f")) |>
  e_tooltip(trigger = "axis") |>
  e_datazoom(x_index = 0) |>
  e_y_axis(name = "USD / barrel") |>
  e_title("WTI vs Brent — daily spot prices", left = "center") |>
  apply_econ_theme()
```

## Brent premium (Brent − WTI) {#spread}

Positive values = Brent trades above WTI.

```{r spread}
#| echo: false
#| eval: !expr have_prices
prices |>
  dplyr::mutate(date = as.Date(date),
                spread = brent - wti) |>
  dplyr::filter(!is.na(spread)) |>
  dplyr::arrange(date) |>
  e_charts(date) |>
  e_line(spread, name = "Brent − WTI ($/bbl)", symbol = "none", smooth = TRUE,
         lineStyle = list(width = 2, color = "#6a4c93"),
         areaStyle = list(color = "rgba(106,76,147,0.15)")) |>
  e_mark_line(data = list(yAxis = 0), lineStyle = list(color = "#adb5bd"),
              symbol = "none") |>
  e_tooltip(trigger = "axis") |>
  e_datazoom(x_index = 0) |>
  e_y_axis(name = "USD / barrel") |>
  e_title("Brent premium over WTI", left = "center") |>
  apply_econ_theme()
```

## Interpretation

- **Wide Brent premium** (>$5) typically signals US shale glut or
  Cushing storage constraints (2011-2014); Middle East supply risk (2018,
  2020 March Saudi-Russia war); or European demand spikes (post-Ukraine).
- **Narrow or inverted spread** (WTI > Brent) is rare and usually reflects
  US-specific supply disruptions or Brent oversupply.
- **Secular narrowing since 2016** — US became a significant crude
  exporter after the 2015 export-ban repeal, arbitraging the two
  benchmarks more tightly.

## What's missing and why

| Data | Why not on this page |
|---|---|
| Country-level retail gasoline / diesel | Not in EIA's free data; IEA publishes behind paid subscription |
| Country-level natural-gas prices | Same — TTF, JKM, NBP benchmarks sit behind commercial feeds |
| Global refining margins | Same |
| Detailed refined-product prices | Same |

A comprehensive international price section would require a paid data feed.
The visualizations shipped here use only public-domain US government data.
