---
title: "Current Fuel Prices — US Map"
subtitle: "PADD-region retail gasoline, most recent weekly snapshot"
---

```{r setup}
#| include: false
source("R/helpers.R")
ENERGY_DIR <- file.path(DATA_ROOT, "energy")

read_csv_safe <- function(p) {
  if (!file.exists(p) || file.size(p) == 0) return(tibble::tibble())
  suppressWarnings(readr::read_csv(p, show_col_types = FALSE))
}

padd <- read_csv_safe(file.path(ENERGY_DIR, "padd_gas_current.csv"))
have <- nrow(padd) > 0

padd_states <- list(
  R10 = c("CT","DE","DC","FL","GA","ME","MD","MA","NH","NJ","NY","NC","PA","RI","SC","VT","VA","WV"),
  R20 = c("IL","IN","IA","KS","KY","MI","MN","MO","NE","ND","OH","OK","SD","TN","WI"),
  R30 = c("AL","AR","LA","MS","NM","TX"),
  R40 = c("CO","ID","MT","UT","WY"),
  R50 = c("AK","AZ","CA","HI","NV","OR","WA")
)
```

::: {.callout-note appearance="simple"}
Retail regular gasoline by **PADD region** (Petroleum Administration for
Defense District). Each state is shaded by its PADD's weekly price.
Source: EIA weekly retail-price survey. Cadence: weekly.
:::

## Current PADD retail gasoline {#map}

```{r map}
#| echo: false
#| eval: !expr have
suppressPackageStartupMessages(library(plotly))

state_df <- purrr::map_dfr(names(padd_states), function(r) {
  row <- padd |> dplyr::filter(duoarea == r)
  if (nrow(row) == 0) return(tibble::tibble(state = padd_states[[r]]))
  tibble::tibble(state = padd_states[[r]],
                 padd  = row$area_name[1],
                 price = row$price_now[1])
}) |>
  dplyr::filter(!is.na(price))

plot_ly(state_df,
        type = "choropleth",
        locationmode = "USA-states",
        locations = ~state,
        z = ~price,
        text = ~paste0("<b>", state, "</b><br>", padd,
                       "<br>Price: $", formatC(price, digits = 3, format = "f"), "/gal"),
        hoverinfo = "text",
        colorscale = list(c(0, "#2f9e44"),
                          c(0.5, "#fcbf49"),
                          c(1, "#bc4749")),
        colorbar = list(title = "$/gal", tickformat = ".2f")) |>
  layout(
    title = list(
      text = paste0("Retail regular gasoline by PADD region — ",
                    format(Sys.Date(), "%B %Y")),
      x = 0.5, xanchor = "center"
    ),
    geo = list(scope = "usa", projection = list(type = "albers usa"),
               showlakes = TRUE, lakecolor = "#ffffff")
  ) |>
  config(displayModeBar = FALSE)
```

## By PADD region {#table}

```{r table}
#| echo: false
#| eval: !expr have
suppressPackageStartupMessages(library(reactable))

reactable(
  padd |>
    dplyr::transmute(padd = area_name, current = price_now,
                     prior_week = price_prior, wow_change),
  striped = TRUE, compact = TRUE, highlight = TRUE,
  theme = reactableTheme(borderColor = "#dee2e6"),
  defaultSorted = list(current = "desc"),
  columns = list(
    padd        = colDef(name = "Region", minWidth = 180),
    current     = colDef(name = "Current ($/gal)",
                         format = colFormat(digits = 3),
                         style = list(fontWeight = 600)),
    prior_week  = colDef(name = "Prior week ($/gal)",
                         format = colFormat(digits = 3)),
    wow_change  = colDef(name = "Δ WoW",
                         format = colFormat(digits = 3),
                         style = function(v) list(
                           color = if (is.na(v)) "#6c757d" else if (v > 0) "#bc4749" else "#2f9e44",
                           fontWeight = 600))
  )
)
```

```{r no-data}
#| echo: false
#| eval: !expr '!have'
htmltools::div(class = "callout callout-warning",
  htmltools::p("PADD gasoline data not yet populated — ensure EIA_API_KEY is ",
               "set and rerun the pipeline."))
```

## Why PADD and not state?

EIA publishes weekly retail gasoline at the **5 PADD regions**
comprehensively. State-level weekly coverage exists for only ~10 states.
For a complete national picture without geographic holes, PADD aggregation
is the honest choice. Every state in a PADD shares the same color because
they share the same weekly aggregate.
